---
title: "EMG modelling"
author: "Johannes Zauner, Lucien Bickerstaff"
date: "`r Sys.Date()`"
output: html_document
---

# Setup

```{r}
# Seed for reproducibility
set.seed(2025)

# Libraries
library(here)
library(tidyverse)
library(LightLogR)
library(patchwork)
library(lme4)
library(visreg)
library(mgcv)
library(sjPlot)
library(broom.mixed)
library(MuMIn)
```

# EMG data

```{r}
# Import EMG activity data
emg_data <- read_csv(here("data", "emg", "emg_pilot_data.csv"))

# create muscle specific response columns
data <- emg_data %>% 
  pivot_wider(names_from = channel,
              values_from = emg_response, # Here we take change counts and not % of change!
              names_prefix = "emg_response_"
              )
```

## Visualize

```{r}
#| warning: false
#| message: false
#| fig-width: 12
#| fig-height: 12

# We take orbicularis oculi response (response_2)
Plot_unscaled <-
data |> 
  ggplot(aes(y = emg_response_2, x = melanopic_edi)) +
  geom_violin(aes(group = melanopic_edi), scale = "width") +
  stat_summary(aes(group = melanopic_edi), fun = mean) +
  stat_summary(aes(group = melanopic_edi), fun = median, col = "red", size = 0.5) +
  labs(title = "Violin plots of response against stimulus with mean (black) and median (red) response\nblue: linear model with 0lx, green: linear model without 0lx") +
  geom_smooth(method = "lm") +
  geom_smooth(aes(group = melanopic_edi == 0), col = "green4", fill = "green4",
              method = "lm")

Plot_scaled <- 
  Plot_unscaled +
  scale_x_continuous(trans = "symlog", breaks = c(0,10^(0:5))) +
  labs(title = NULL)

Plot_unscaled_log10 <-
  Plot_unscaled +
    scale_y_log10() +
  labs(title = NULL, subtitle = "Y scaled log10")
  
Plot_scaled_log10 <- 
  Plot_scaled +
  scale_y_log10()

(Plot_unscaled + Plot_scaled) /
(Plot_unscaled_log10 + Plot_scaled_log10)
```

## Model

Including random intercept for pid

```{r}
#| fig-width: 8
#| fig-height: 12

data <- 
data |> 
  mutate(is.zero = melanopic_edi == 0,
         log_mel_edi = pmax(melanopic_edi, 1) |> log10())

m1_formula <- log10(emg_response_2) ~ is.zero + log_mel_edi + (1| participant_id)
m0_formula <- log10(emg_response_2) ~ log_mel_edi + (1| participant_id)
        
model1 <- lmer(m1_formula, data   = data)
model0 <- lmer(m0_formula, data   = data)

summary(model1)
summary(model0)

performance::check_model(model1)
performance::check_model(model0)
anova(model0, model1)
```

```{r}
visreg(model1, jitter = TRUE)
```

Not a perfect fit, but good enough diagnostics. Better than the Gamma model.

### Checking whether the dependency is log linear around the data

This is done to see whether if we free the modelling approach from the linearity assumption, does it capture non-linear behaviour that we might have overlooked?

```{r}
data$log_mel_shift <- log(data$melanopic_edi + 1)

m_gamm <- gam(
  emg_response_2 ~ s(log_mel_shift, k = 6) +
    s(participant_id, bs = "re"),
  family = Gamma(link = "log"),
  data = data
)

summary(m_gamm)
plot(m_gamm, all.terms = TRUE)
```

Looking good:

That is a generalized additive mixed model. It does not assume a linear relationship between the variables and thus is useful if we want to assess whether a non-linear relationship agrees with our linear model - in the relevant region, that is. So if we only look to the right of the plot - that is where the non-zero Mel EDI data are. And there we don’t see relevant non-linearities, so our approach is not missing anything important.
You can see where our data are by the small lines sticking up from the x-axis

# Saving final model

The model including is.zero is a better fit. We drop participant-dependent slopes as they prevent models from converging.

```{r}
saveRDS(model1, here("outputs", "emg_lmm.rds"))
```

# Manuscript-ready outputs

```{r}
tab_model(
  model1,
  show.re.var = TRUE,   # show random-effect SD/variance
  show.icc = TRUE,      # intraclass correlation
  show.r2 = TRUE,       # pseudo-R² if available
  pred.labels = c("Intercept", "is.zeroTRUE", "log_mel_edi"),
  dv.labels = "emg_response",
  file = here("outputs", "model_table.html")
)

coef_table <- tidy(model1, effects = "fixed", conf.int = TRUE)
coef_table

r2 <- r.squaredGLMM(model1)
r2
```

```{r}
sessionInfo()
```
