---
title: "Compute primary settings"
author: "Lucien Bickerstaff"
date: "`r Sys.Date()`"
output: html_document
---

# Setup

```{r}
# Seed for reproducibility
set.seed(2025)

# Libraries
library(here)
library(tidyverse)
library(openxlsx)
library(photobiology)
library(grid)
library(patchwork)
library(glue)

source(here("R", "plot_settings.R"))
```

# Read calibration data

```{r warning=F}
# Spectral measurements done using the Jeti (with screen), for each LED and for 
# white light (all LED at same intensity setting), at these settings: 
# 25, 50, 100, 200, 400, 800, 1200, 1600, 2000, 2400, 2800, 3200, 3600, 4000, 4095/4095
data <- read.xlsx(here("data", "20240716_stlab_spectral_calibration_measurements_fullexport.xlsx"),
                  colNames = F)
# Transpose the data
data <- as.data.frame(t(data), stringsAsFactors = FALSE)
# Assign the first row as header
colnames(data) <- data[1, ]
data <- data[-1, , drop = FALSE]
```

# Extract mel EDI and photopic lux calibration information

```{r warning=F, echo=TRUE, fold=FALSE}
lux_data <- data %>% 
  select(Name, `Ev [lx] (CIE1931 2∞)`, `X v,mel,D65 (MEDI) [lx]`) %>% 
  mutate(
    measurement = str_extract(Name, "(?<=#)\\d+"),
    measurement = factor(measurement)
  ) %>% 
  mutate(measurement = fct_reorder(measurement, as.numeric(as.character(measurement)))) %>% 
  select(-Name)

lux_data <- lux_data %>%
  mutate(
    measurement_num = as.numeric(measurement),
    across(
      where(is.character),
      ~ as.numeric(gsub(",", ".", .))  # convert European decimal commas
    )
  )
# This converts all character columns (like "1,748E-01") into proper numeric values

glimpse(lux_data)
```

# Calculate total irradiance and merge calibration data

## Extract spectral data

```{r warning=F, echo=TRUE, fold=FALSE}
spectral_data <- data %>% 
  select(c(Name, "380":"780"))

spectral_data <- spectral_data %>%
  pivot_longer(
    cols = -1,                        # all columns except the first
    names_to = "wavelength",          # column for wavelength names
    values_to = "Ee [W/(sqm*nm)]"           # column for measured values
  ) %>%
  mutate(across(  # ensure numeric x-axis
    -1,  # skip the first column (the names)
    ~ as.numeric(gsub(",", ".", .))
  )) %>% 
  mutate(
    measurement = str_extract(Name, "(?<=#)\\d+"),
    measurement = factor(measurement)
  ) %>% 
  mutate(measurement = fct_reorder(measurement, as.numeric(as.character(measurement)))) %>% 
  select(-Name)
```

### Control plot…

```{r}
gradient <- t(w_length2rgb(400:700))
g <- rasterGrob(gradient, width = unit(1, "npc"), height = unit(1, "npc"), interpolate = TRUE) 

ggplot(spectral_data %>% filter(measurement == 15), 
       aes(x = wavelength, y = `Ee [W/(sqm*nm)]`)) +
  geom_line(alpha = 0.7, linewidth = 0.8) +
  geom_hline(yintercept = 0, color = "darkred") +
  annotation_custom(g, xmin = -Inf, xmax = Inf, ymin = -0.1, ymax = -0.05) +
  labs(
    x = "Wavelength (nm)",
    y = "Ee [W/(m²·nm)]"
  ) +
  ylim(-0.1, 1) +
  theme_minimal() +
  theme(
    legend.position = "right"
  ) +
  plot_formatting
```

We calculate total irradiance as the sum of the irradiances at all wavelengths

```{r warning=F, message=F, echo=TRUE, fold=FALSE}
irradiance_data <- spectral_data %>% 
  group_by(measurement) %>%
  summarise(`Ee [W/(sqm)]` = sum(`Ee [W/(sqm*nm)]`), .groups = "drop")

irradiance_data <- irradiance_data %>%
  mutate(measurement_num = as.numeric(as.character(measurement)))

merged_data <- left_join(
  irradiance_data, 
  lux_data, by = c("measurement", "measurement_num")
  )

merged_data <- merged_data %>%
  mutate(
    primary = case_when(
      measurement_num >= 1   & measurement_num <= 15  ~ "1",
      measurement_num >= 16  & measurement_num <= 30  ~ "2",
      measurement_num >= 31  & measurement_num <= 45  ~ "3",
      measurement_num >= 46  & measurement_num <= 60  ~ "4",
      measurement_num >= 61  & measurement_num <= 75  ~ "5",
      measurement_num >= 76  & measurement_num <= 90  ~ "6",
      measurement_num >= 91  & measurement_num <= 105 ~ "7",
      measurement_num >= 106 & measurement_num <= 120 ~ "8",
      measurement_num >= 121 & measurement_num <= 135 ~ "9",
      measurement_num >= 136 & measurement_num <= 150 ~ "10",
      measurement_num >= 151 & measurement_num <= 165 ~ "all",
      TRUE ~ NA_character_    # for the final "all off" (166)
    ),
    primary = factor(primary, levels = c(as.character(1:10), "all"))
  )

merged_data <- merged_data %>%
  filter(!is.na(primary))

merged_data <- merged_data %>%
  group_by(primary) %>%
  mutate(meas_within_primary = row_number()) %>%
  ungroup()

# map settings to the measurements
settings_vec <- c(25, 50, 100, 200, 400, 800, 1200, 1600, 2000, 
                  2400, 2800, 3200, 3600, 4000, 4095)
merged_data <- merged_data %>%
  mutate(setting = settings_vec[meas_within_primary])
```

## Control plots…

```{r}
ggplot(data=merged_data, aes(x=setting, y=`Ee [W/(sqm)]`)) +
  geom_point() +
  labs(title = "Irradiance") +
  facet_wrap(~ primary) +
  plot_formatting
ggplot(data=merged_data, aes(x=setting, y=`Ev [lx] (CIE1931 2∞)`)) +
  geom_point() +
  labs(title = "Photopic illuminance") +
  facet_wrap(~ primary) +
  plot_formatting
ggplot(data=merged_data, aes(x=setting, y=`X v,mel,D65 (MEDI) [lx]`)) +
  geom_point() +
  labs(title = "Melanopic EDI") +
  facet_wrap(~ primary) +
  plot_formatting
```

# Find max possible irradiance for each primary of interest

```{r echo=TRUE, fold=FALSE}
merged_data_select <- merged_data %>%
  filter(primary %in% as.character(2:7))

max_irradiance_data <- merged_data_select %>%
  group_by(primary) %>%
  summarise(
    max_irr = max(`Ee [W/(sqm)]`, na.rm = TRUE)
  )

# The max possible irradiance value is the min irradiance value achievable
# by all primaries overall
min_irradiance <- min(max_irradiance_data$max_irr)
min_irradiance
```

This is the maxmimum irradiance we can achieve across all primaries.

# Define 4 log-spaced target irradiances 

Four target irradiance values for each of the four lab visits

```{r echo=TRUE, fold=FALSE}
decrease_factor <- 150 # how much smaller the min irradiance value is compared to the max value

max_val <- min_irradiance
min_val <- max_val / decrease_factor
n_values <- 4

irradiance_levels <- exp(seq(log(max_val), log(min_val), length.out = n_values))
irradiance_levels
```

# Interpolation of calibration data

We interpolate the data so we have a [setting] -> [intensity metric] map

```{r echo=TRUE, fold=FALSE}
# Define target settings (full interpolation range)
all_settings <- 0:4095  # start from 0 now

interp_data <- merged_data %>%
  group_by(primary) %>%
  group_modify(~ {
    df <- .x[order(.x$setting), ]

    # Base interpolation range (0 to 4095)
    xout <- all_settings

    # --- Irradiance ---
    Ee_fit <- spline(
      x = df$setting,
      y = df$`Ee [W/(sqm)]`,
      xout = xout
    )$y

    # --- Photopic illuminance ---
    Ev_fit <- spline(
      x = df$setting,
      y = df$`Ev [lx] (CIE1931 2∞)`,
      xout = xout
    )$y

    # --- Melanopic EDI ---
    MEDI_fit <- spline(
      x = df$setting,
      y = df$`X v,mel,D65 (MEDI) [lx]`,
      xout = xout
    )$y

    # --- Linear extension to 0 ---
    # Extrapolate linearly from first data point back to 0
    first_x <- df$setting[1]
    first_Ee <- df$`Ee [W/(sqm)]`[1]
    first_Ev <- df$`Ev [lx] (CIE1931 2∞)`[1]
    first_MEDI <- df$`X v,mel,D65 (MEDI) [lx]`[1]

    # Fill any values below the first setting (should be 0–first_x)
    Ee_fit[xout < first_x] <- (xout[xout < first_x] / first_x) * first_Ee
    Ev_fit[xout < first_x] <- (xout[xout < first_x] / first_x) * first_Ev
    MEDI_fit[xout < first_x] <- (xout[xout < first_x] / first_x) * first_MEDI

    tibble(
      setting = xout,
      Ee = Ee_fit,
      Ev = Ev_fit,
      MEDI = MEDI_fit
    )
  }) %>%
  ungroup()
```

## Control plots…

```{r}
# Irradiance
ggplot(interp_data, aes(x = setting, y = Ee, color = primary)) +
  geom_line() +
  geom_point(data = merged_data, aes(y = `Ee [W/(sqm)]`), color = "black", size = 0.8) +
  theme_minimal() +
  facet_wrap(~ primary, scales = "free") +
  labs(y = "Irradiance (W/m²)", title = "Interpolated Irradiance per Primary") +
  plot_formatting

# Illuminance
ggplot(interp_data, aes(x = setting, y = Ev, color = primary)) +
  geom_line() +
  geom_point(data = merged_data, aes(y = `Ev [lx] (CIE1931 2∞)`), color = "black", size = 0.8) +
  theme_minimal() +
  facet_wrap(~ primary, scales = "free") +
  labs(y = "Ev [lx] (CIE1931 2°)", title = "Interpolated Photopic Illuminance per Primary") +
  plot_formatting

# Melanopic EDI
ggplot(interp_data, aes(x = setting, y = MEDI, color = primary)) +
  geom_line() +
  geom_point(data = merged_data, aes(y = `X v,mel,D65 (MEDI) [lx]`), color = "black", size = 0.8) +
  theme_minimal() +
  facet_wrap(~ primary, scales = "free") +
  labs(y = "X v,mel,D65 (MEDI) [lx]", title = "Interpolated Melanopic EDI per Primary") +
  plot_formatting
```

# Map target irradiance levels to needed settings for each primary

Now, we just map our target irradiance levels to the desired settings…

```{r warning=F, echo=TRUE, fold=FALSE}
interp_data_all <- interp_data # keeping old df…

interp_data <- interp_data %>% 
  filter(primary != "all") %>%
  filter(primary %in% as.character(2:7))

# For each primary and each target irradiance, find the closest setting
mapped_settings <- expand_grid(
  primary = unique(interp_data$primary),
  target_irr = irradiance_levels
) %>%
  left_join(interp_data, by = "primary") %>%
  group_by(primary, target_irr) %>%
  summarise(
    # Find index of nearest irradiance to the target
    idx = which.min(abs(Ee - target_irr)),

    # Extract corresponding setting and all interpolated modalities
    setting = setting[idx],
    calculated_Ee = Ee[idx],
    calculated_Ev = Ev[idx],
    calculated_MEDI = MEDI[idx],
    .groups = "drop"
  ) %>%
  group_by(primary) %>%
  mutate(
    lab_visit = dense_rank(target_irr),  # 1 = lowest, 4 = highest
    lab_visit = factor(lab_visit, levels = 1:4, labels = c("1", "2", "3", "4"))
  ) %>%
  ungroup() %>% 
  select(-idx)
```

# Add peak wavelength information

Extracted from Spectra Tune Lab manual

```{r}
peak_nm_df <- data.frame(primary = factor(c(1:10)),
                         peak_nm = c(430, 445, 465, 475, 505, 525, 545, 595, 640, 660),
                         color = c("UV", "Royal blue", "Dark blue", "Blue", "Cyan", "Green", "Lime", "PC Amber", "Red", "Deep red"))

mapped_settings <- mapped_settings %>% 
  left_join(peak_nm_df, by = "primary")
mapped_settings
```

# Calculate coverage

```{r}
irradiance_levels_log10 <- log10(mapped_settings$calculated_Ee)
coverage <- max(irradiance_levels_log10) - min(irradiance_levels_log10)
coverage
```

# Check settings…

Here we did calibration measurements to check whether the calculated settings are as expected in terms of output

```{r fig.width=5, fig.height=3}
# calculated settings and theoretical irradiance, illuminance and mel EDI 
theoretical <- mapped_settings

# calibration data: true irradiance, illuminance and mel EDI
measured <- read.xlsx(here("data", "20251028_new_settings_all.xlsx"),
                      colNames = F)

process_spectral_measurements <- function(measured) {
  # Transpose the data
  measured <- as.data.frame(t(measured), stringsAsFactors = FALSE)
  # Assign the first row as header
  colnames(measured) <- measured[1, ]
  measured <- measured[-1, , drop = FALSE]
  
  measured <- measured %>% 
    select(Name, `Irradiance [W/sqm] (380-780nm)`, `Illuminance [lx] (CIE1931 2°)`, `X v,mel,D65 (MEDI) [lx]`) %>% 
    mutate(
      measurement = str_extract(Name, "(?<=#)\\d+"),
      measurement = factor(measurement)
    ) %>% 
    mutate(measurement = fct_reorder(measurement, as.numeric(as.character(measurement)))) %>% 
    select(-Name)
  
  measured <- measured %>%
    mutate(
      measurement_num = as.numeric(measurement),
      across(
        where(is.character),
        ~ as.numeric(gsub(",", ".", .))  # convert European decimal commas
      )
    )
  # This converts all character columns (like "1,748E-01") into proper numeric values
  
  measured <- measured %>%
    mutate(
      # Identify which repetition (1–6)
      repetition = ceiling(measurement_num / (6 * 4)),
  
      # Index within repetition (1–24)
      within_rep = measurement_num - (repetition - 1) * 24,
  
      # Primary assignment (2–7)
      primary_num = case_when(
        repetition <= 3 ~ ceiling(within_rep / 4) + 1,        # forward: 2–7
        repetition > 3  ~ 8 - ceiling(within_rep / 4)         # reverse: 7–2
      ),
  
      # Lab visit number 1–4, clean modulo form
      lab_visit_num = ((within_rep - 1) %% 4) + 1,
  
      # Reverse lab visit order for reverse reps
      lab_visit = case_when(
        repetition <= 3 ~ as.character(lab_visit_num),         # forward: 1–4
        repetition > 3  ~ as.character(5 - lab_visit_num)      # reverse: 4–1
      ),
  
      # Final labeling
      primary = factor(primary_num, levels = as.character(2:7)),
      lab_visit = factor(lab_visit, levels = as.character(1:4))
    )

  plot <- measured %>% 
    ggplot(aes(x = primary, y = `Irradiance [W/sqm] (380-780nm)`, color = lab_visit)) +
    geom_point() +
    scale_y_log10() +
    theme_minimal() +
    plot_formatting
  
  print(plot)
  
  measured_means <- measured %>% 
    group_by(primary, lab_visit) %>% 
    summarise(`Irradiance [W/sqm] (380-780nm)` = mean(`Irradiance [W/sqm] (380-780nm)`, na.rm = T),
              `Illuminance [lx] (CIE1931 2°)` = mean(`Illuminance [lx] (CIE1931 2°)`, na.rm = T),
              `X v,mel,D65 (MEDI) [lx]` = mean(`X v,mel,D65 (MEDI) [lx]`, na.rm. =T))
  
  merged <- measured_means %>% 
    left_join(theoretical, by = c("primary", "lab_visit"))
  
  ## Calculate melanopic to photopic ratios
  merged <- merged %>% 
    mutate(mel_photopic_ratio = `X v,mel,D65 (MEDI) [lx]`/`Illuminance [lx] (CIE1931 2°)`)
  
  return(merged)
}

merged <- process_spectral_measurements(measured)
```

# Visualise differences between calculated and measured 

```{r}
plot_diff_measured <- function(df, save=F) {
  
  # Irradiance
  plot_ee <- df %>% 
    ggplot(aes(x = calculated_Ee, y = `Irradiance [W/sqm] (380-780nm)`, colour = lab_visit)) +
    geom_point() +
    geom_abline(intercept = 0, slope = 1, color = "black") +
    facet_wrap(~ primary) +
    scale_y_log10() +
    scale_x_log10() +
    labs(x = "Calculated (log10 scale)",
         y = "Measured (log10 scale)",
         title = "Irradiance [W/sqm] (380-780nm)") +
    theme_minimal() +
    plot_formatting

  # Illuminance
  plot_ev <- df %>% 
    ggplot(aes(x = calculated_Ev, y = `Illuminance [lx] (CIE1931 2°)`, colour = lab_visit)) +
    geom_point() +
    geom_abline(intercept = 0, slope = 1, color = "black") +
    facet_wrap(~ primary) +
    scale_y_log10() +
    scale_x_log10() +
    labs(x = "Calculated (log10 scale)",
         y = "Measured (log10 scale)",
         title = "Illuminance [lx]") +
    theme_minimal() +
    plot_formatting
  
  # Melanopic EDI
  plot_medi <- df %>% 
    ggplot(aes(x = calculated_MEDI, y = `X v,mel,D65 (MEDI) [lx]`, colour = lab_visit)) +
    geom_point() +
    geom_abline(intercept = 0, slope = 1, color = "black") +
    facet_wrap(~ primary) +
    scale_y_log10() +
    scale_x_log10() +
    labs(x = "Calculated (log10 scale)",
         y = "Measured (log10 scale)",
         title = "Melanopic EDI [lx]") +
    theme_minimal() +
    plot_formatting
  
  print(plot_ee)
  print(plot_ev)
  print(plot_medi)
  
  if (save) {
    formatted_save(here("outputs", "figures", "calibration_ee_accuracy.jpg"), plot_ee)
    formatted_save(here("outputs", "figures", "calibration_ee_accuracy.jpg"), plot_ev)
    formatted_save(here("outputs", "figures", "calibration_medi_accuracy.jpg"), plot_medi)
  }
}
``` 

```{r fig.width=5, fig.height=3}
plot_diff_measured(merged)
```

# Visualise differences in measured metrics between primaries

```{r}
plot_diff_metrics <- function(df, save=F) {
  # Irradiance
  plot_ee <- df %>% 
    ggplot(aes(x = primary, 
               y = `Irradiance [W/sqm] (380-780nm)`, 
               colour = lab_visit,
               group = lab_visit)) +
    geom_point() +
    geom_line() +
    scale_y_log10() +
    scale_colour_manual(values = colorRampPalette(c("gold", "red3"))(4)) +
    labs(title = "Irradiance",
         y = "Irradiance [W/sqm] (380-780nm)",
         x = "Primary",
         colour = "Light intensity") +
    theme_minimal() +
    plot_formatting
  
  # Illuminance
  plot_ev <- df %>% 
    ggplot(aes(x = primary, 
               y = `Illuminance [lx] (CIE1931 2°)`, 
               colour = lab_visit,
               group = lab_visit)) +
    geom_point() +
    geom_line() +
    scale_y_log10() +
    labs(title = "Illuminance",
         y = "Illuminance [lx]",
         x = "Primary",
         colour = "Light intensity") +
    theme_minimal() +
    plot_formatting
  
  # Melanopic EDI
  plot_medi <- df %>% 
    ggplot(aes(x = primary, 
               y = `X v,mel,D65 (MEDI) [lx]`, 
               colour = lab_visit, 
               group = lab_visit)) +
    geom_point() +
    geom_line() +
    scale_y_log10() +
    labs(title = "Melanopic EDI",
         y = "Melanopic EDI [lx]",
         x = "Primary",
         colour = "Light intensity") +
    theme_minimal() +
    plot_formatting
  
  # Photopic vs Melanopic
  df_long <- df %>%
    select(primary, lab_visit, 
           Melanopic = `X v,mel,D65 (MEDI) [lx]`, 
           Photopic  = `Illuminance [lx] (CIE1931 2°)`) %>%
    pivot_longer(cols = c(Melanopic, Photopic),
                 names_to = "type",
                 values_to = "value")
  
  plot_ratio_1 <- ggplot(df_long, # %>% mutate(primary = as.factor(as.numeric(primary))), 
                         aes(x = primary, 
                             y = value, 
                             shape = type,
                             linetype = type,
                             colour = lab_visit,
                             group = interaction(lab_visit, type))) +
    geom_point() + 
    geom_line() +
    scale_y_log10() +
    scale_colour_manual(values = colorRampPalette(c("gold", "red3"))(4)) +
    labs(title = "Melanopic vs Photopic",
         y = "Illuminance [lx]",
         x = "Primary",
         colour = "Light intensity",
         linetype = "Type",
         shape = "Type") +
    theme_minimal() +
    theme(legend.key.height = unit(0.4, "cm")) +
    plot_formatting
  
  # Melanopic vs photopic ratios
  plot_ratio_2 <- df %>% 
    ggplot(aes(x = primary, y = mel_photopic_ratio, colour = lab_visit)) + 
    geom_hline(yintercept = 1, colour = "darkred") +
    # geom_text(label = "1:1 ratio", vjust = -1) +
    geom_point() +
    # geom_smooth() +
    scale_y_log10() +
    labs(title = "Melanopic vs Photopic",
         y = "Melanopic/Photopic lux ratio (log10 scale)",
         x = "Primary",
         colour = "Light intensity") +
    theme_minimal() +
    plot_formatting
  
  print(plot_ee)
  print(plot_ev)
  print(plot_medi)
  print(plot_ratio_1)
  print(plot_ratio_2)
  
  if (save) {
    saveRDS(plot_ee, here("outputs", "figures", "rds", glue("calibration_ee_metrics.rds")))
    saveRDS(plot_ratio_1, here("outputs", "figures", "rds", glue("calibration_ee_medi_ratio_1.rds")))
    
    formatted_save(here("outputs", "figures", "calibration_ee_metrics.jpg"), plot_ee)
    formatted_save(here("outputs", "figures", "calibration_ev_metrics.jpg"), plot_ev)
    formatted_save(here("outputs", "figures", "calibration_medi_metrics.jpg"), plot_medi)
    formatted_save(here("outputs", "figures", "calibration_ee_medi_ratio_1.jpg"), plot_ratio_1)
    formatted_save(here("outputs", "figures", "calibration_ee_medi_ratio_2.jpg"), plot_ratio_2)
  }
}
```

```{r fig.width=5, fig.height=3}
plot_diff_metrics(merged)
```

We see that for the second lab visit, the irradiance is not matched…
So, we manually correct it and check with new spectral measurements

```{r fig.width=5, fig.height=3}
final_settings <- merged %>% 
  select(-c(`Irradiance [W/sqm] (380-780nm)`,
            `Illuminance [lx] (CIE1931 2°)`,
            `X v,mel,D65 (MEDI) [lx]`,
            calculated_Ee,
            calculated_Ev,
            calculated_MEDI))

spec_data_updated <- read.xlsx(here("data", "20251105_corrected_settings.xlsx"),
                               colNames = F)

spec_data_updated <- process_spectral_measurements(spec_data_updated)

# Update the settings from the old DF that have been manually adjusted
spec_data_updated <- spec_data_updated %>% 
  mutate(
    setting = case_when(lab_visit == 2 & primary == 2 ~ 128,
                        lab_visit == 2 & primary == 3 ~ 142,
                        lab_visit == 2 & primary == 4 ~ 156,
                        lab_visit == 2 & primary == 5 ~ 166,
                        lab_visit == 2 & primary == 6 ~ 174,
                        lab_visit == 2 & primary == 7 ~ 115,
                        TRUE ~ setting)
)

plot_diff_measured(spec_data_updated, save = T)
plot_diff_metrics(spec_data_updated, save = T)
```

There is a discrepancy between what is calculated initially and what is true.
Main thing is that the irradiance is constant across primaries, and well log-spaced between lab visits.

# Save design matrix

```{r}
write.csv2(spec_data_updated, here("outputs", "design_matrix.csv"), row.names = F)
```

```{r}
sessionInfo()
```
