---
title: "Power analysis"
author: "Lucien Bickerstaff"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
---

```{r render-time-1, include=FALSE}
# Record how long it took to render the file
.render_start <- Sys.time()
```

```{r knitr-options, include=FALSE}
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE
)
```

# Setup

```{r setup, warning=FALSE, message=FALSE}
# Seed for reproducibility
seed <- 2025
set.seed(seed)

# Libraries
library(here)
library(glue)
library(broom.mixed)
library(jsonlite)
library(BayesFactor)
library(lme4)
library(glmmTMB)
library(MASS)
library(tidyverse)
library(furrr)
library(patchwork)
library(ggtext)
library(progressr)

# Paths
source(here("R", "helpers.R"))
source(here("R", "plot_settings.R"))
```

A few notes:

- We define a function to simulate data from our fitted models
- We simulate and visualise the data
- Because the design matrix extends to lower light levels than what we had
during the pilot study, we have to use extrapolation. However, this means that
simulated data might span values that are non-physiological, e.g. lower than
what we observed for the 0-lx condition in the pilot study. So, we tried:
1. Flooring non-physiological data. This introduces a non-linear component that 
is undesirable as it makes convergence harder and may not represent the actual 
responses, so we did not choose this option
2. Discarding non-physiological data. This is the safer option as it does not
assume anything about the actual responses.
- We then run a power analysis:
1. Consider a range of sample sizes and residual noise scaling levels for 
simulation
2. Compare melanopic vs photopic models with AIC-based model comparison
3. Look at proportions of comparisons favouring one, the other, or no model 
4. Simulate 1000 datasets per comparison

# Load and prepare data

## Load models

We use OO-EMG as our reference modality, as it is the most reliable and the 
most indicative based on literature (see manuscript)

```{r load-driver-models}
m_mel <- readRDS(here("outputs", "models", "emg_lmm_mel.rds"))
m_phot <- readRDS(here("outputs", "models", "emg_lmm_phot.rds"))
m_rho <- readRDS(here("outputs", "models", "emg_lmm_rho.rds"))

driver_models <- tibble::tibble(
  driver = c("mel", "phot", "rho"),
  model = list(m_mel, m_phot, m_rho)
)
```

## Load design matrix

…from `generate_design_matrix.Rmd`

```{r load-design-matrix}
design_matrix <- readr::read_csv(
  here::here("outputs", "design_matrix.csv"),
  show_col_types = FALSE
) %>%
  dplyr::mutate(
    participant_id = factor(participant_id),
    lab_visit = factor(lab_visit),

    melanopic_edi = MEDI,
    rhodopic_edi = REDI,

    # Must match the transformation used when fitting the EMG model.
    # Even though there is no 0-lx condition here
    log_illuminance = log10(pmax(illuminance, 1)),
    log_melanopic_edi = log10(pmax(melanopic_edi, 1)),
    log_rhodopic_edi  = log10(pmax(rhodopic_edi, 1)),

    lights_on = melanopic_edi != 0
  )
```

Simulation matrix also needs `log_baseline`.

All three models were fitted to the same EMG data, so `log_baseline` is identical 
across model frames. So we just use one model, for example the melanopic model.

We first check it though:

```{r check-baseline-model-frames}
baseline_check <- driver_models %>%
  dplyr::mutate(
    log_baseline = purrr::map(model, ~ model.frame(.x)$log_baseline)
  )

identical(
  baseline_check$log_baseline[[1]],
  baseline_check$log_baseline[[2]]
)

identical(
  baseline_check$log_baseline[[1]],
  baseline_check$log_baseline[[3]]
)
```

All identical as expected.
We store `log_baseline` values, so we can resample from them during simulation

```{r store-log-baseline}
# -----------------------------
# Store observed log_baseline values for empirical resampling
# -----------------------------

model_frame <- m_mel$frame

observed_log_baseline <- model_frame$log_baseline
observed_log_baseline <- observed_log_baseline[is.finite(observed_log_baseline)]

if (length(observed_log_baseline) == 0) {
  stop(
    "No finite observed log_baseline values available for resampling.",
    call. = FALSE
  )
}

# Quick visual check of the empirical baseline distribution
tibble::tibble(
  log_baseline = observed_log_baseline
) %>%
  ggplot2::ggplot(
    ggplot2::aes(x = log_baseline)
  ) +
  ggplot2::geom_histogram(bins = 30, colour = "white") +
  ggplot2::theme_classic() +
  ggplot2::labs(
    x = "Observed log_baseline",
    y = "Count",
    title = "Empirical distribution used for baseline resampling"
  )
```

```{r}
# -----------------------------
# Simulate log_baseline by empirical resampling
# -----------------------------

model_frame <- m_mel$frame

observed_log_baseline <- model_frame$log_baseline
observed_log_baseline <- observed_log_baseline[is.finite(observed_log_baseline)]

if (length(observed_log_baseline) == 0) {
  stop("No finite observed log_baseline values available for resampling.", call. = FALSE)
}

# Resample one baseline value per planned participant.
# This keeps log_baseline constant within participant, which is usually more
# appropriate than drawing a new baseline for every row/trial.
baseline_by_participant <- design_matrix %>%
  dplyr::distinct(participant_id) %>%
  dplyr::mutate(
    log_baseline = sample(
      observed_log_baseline,
      size = dplyr::n(),
      replace = TRUE
    )
  )

design_matrix <- design_matrix %>%
  dplyr::select(-dplyr::any_of("log_baseline")) %>%
  dplyr::left_join(
    baseline_by_participant,
    by = "participant_id"
  )

# Observed vs simulated
observed_plot_df <- tibble::tibble(
  source = "Observed",
  log_baseline = observed_log_baseline
)

simulated_plot_df <- baseline_by_participant %>%
  dplyr::select(log_baseline) %>%
  dplyr::mutate(source = "Simulated")

plot_df <- dplyr::bind_rows(observed_plot_df, simulated_plot_df)

ggplot(plot_df, aes(x = log_baseline)) +
  geom_histogram(bins = 30, colour = "white") +
  facet_wrap(~ source, scales = "free_y") +
  theme_classic() +
  labs(
    x = "log_baseline",
    y = "Count",
    title = "Observed vs empirically resampled log_baseline"
  )
```


# Define functions

## Data simulation

```{r define-simulation-function}
simulate_from_design <- function(design_df,
                                 model,
                                 sd_scaling = 1,
                                 seed = seed,
                                 baseline_pool = NULL) {
  if (!is.null(seed)) {
    set.seed(seed)
  }

  if (!requireNamespace("MASS", quietly = TRUE)) {
    stop("Package 'MASS' is required.", call. = FALSE)
  }

  if (!requireNamespace("lme4", quietly = TRUE)) {
    stop("Package 'lme4' is required.", call. = FALSE)
  }

  if (!requireNamespace("glmmTMB", quietly = TRUE)) {
    stop("Package 'glmmTMB' is required.", call. = FALSE)
  }

  # ---------------------------------------------------------------------------
  # Prepare and check design matrix
  # ---------------------------------------------------------------------------

  dat <- design_df %>%
    dplyr::mutate(
      participant_id = factor(participant_id),
      trial = as.numeric(trial),
      lights_on = as.logical(lights_on)
    )
  
  # Optionally resample log_baseline for every row/trial.
  # Because this happens inside simulate_from_design(), each simulation replicate
  # receives a fresh set of baseline values.
  if (!is.null(baseline_pool)) {
    
    baseline_pool <- baseline_pool[is.finite(baseline_pool)]
    
    if (length(baseline_pool) == 0) {
      stop(
        "`baseline_pool` contains no finite values.",
        call. = FALSE
      )
    }
    
    dat <- dat %>%
      dplyr::mutate(
        log_baseline = sample(
          baseline_pool,
          size = dplyr::n(),
          replace = TRUE
        )
      )
  }

  # Variables needed by the model RHS, including random-effect grouping variables
  required_vars <- all.vars(stats::formula(model)[[3]])

  missing_vars <- setdiff(required_vars, names(dat))

  if (length(missing_vars) > 0) {
    stop(
      "The design matrix is missing variables required by the model: ",
      paste(missing_vars, collapse = ", "),
      call. = FALSE
    )
  }

  # Check for missing predictor values
  predictor_na_counts <- dat %>%
    dplyr::select(dplyr::all_of(required_vars)) %>%
    dplyr::summarise(
      dplyr::across(
        dplyr::everything(),
        ~ sum(is.na(.x))
      )
    )

  if (any(predictor_na_counts > 0)) {
    print(predictor_na_counts)
    stop(
      "Some model predictors contain NA values in the design matrix.",
      call. = FALSE
    )
  }

  # ---------------------------------------------------------------------------
  # Fixed-effect prediction
  # ---------------------------------------------------------------------------

  # Fixed-effects-only prediction on the model scale.
  # For a Gaussian identity-link model fitted to log_emg_raw, this is the
  # predicted log_emg_raw value without participant-specific random effects.
  dat$eta_fixed <- predict(
    model,
    newdata = dat,
    re.form = NA,
    type = "link",
    allow.new.levels = TRUE
  )

  # ---------------------------------------------------------------------------
  # Simulate random effects
  # ---------------------------------------------------------------------------

  dat$eta_random <- 0

  random_terms <- lme4::findbars(stats::formula(model))

  vc <- glmmTMB::VarCorr(model)$cond

  for (term in random_terms) {

    random_effect_formula <- term[[2]]
    grouping_factor_expr  <- term[[3]]

    grouping_vars <- all.vars(grouping_factor_expr)
    grouping_name <- paste(grouping_vars, collapse = ":")

    if (!grouping_name %in% names(vc)) {
      warning(
        "Random-effect grouping factor '", grouping_name,
        "' not found in VarCorr(model). Skipping this random effect."
      )
      next
    }

    vc_matrix <- as.matrix(vc[[grouping_name]])

    # Examples:
    #   (1 | participant_id) becomes ~ 1
    #   (log_melanopic_edi | participant_id) becomes ~ log_melanopic_edi
    # Because model.matrix(~ x) includes the intercept, this correctly gives
    # random intercept + random slope for (x | participant_id).
    re_formula <- stats::as.formula(
      paste("~", paste(deparse(random_effect_formula), collapse = ""))
    )

    Z <- stats::model.matrix(re_formula, data = dat)

    group_id <- if (length(grouping_vars) == 1) {
      dat[[grouping_vars]]
    } else {
      interaction(dat[grouping_vars], drop = TRUE)
    }

    group_id <- factor(group_id)

    simulated_re <- MASS::mvrnorm(
      n = nlevels(group_id),
      mu = rep(0, ncol(vc_matrix)),
      Sigma = vc_matrix
    )

    simulated_re <- as.matrix(simulated_re)

    colnames(simulated_re) <- colnames(vc_matrix)
    rownames(simulated_re) <- levels(group_id)

    common_cols <- intersect(colnames(Z), colnames(simulated_re))

    if (length(common_cols) == 0) {
      warning(
        "No matching random-effect columns for grouping factor '",
        grouping_name,
        "'. Skipping this random effect."
      )
      next
    }

    re_contribution <- rowSums(
      Z[, common_cols, drop = FALSE] *
        simulated_re[as.character(group_id), common_cols, drop = FALSE]
    )

    dat$eta_random <- dat$eta_random + re_contribution
  }

  # ---------------------------------------------------------------------------
  # Simulate residual noise and back-transform response
  # Floor is applied to the expected response before residual noise is added
  # ---------------------------------------------------------------------------

  sd_resid <- stats::sigma(model) * sd_scaling

  # Fixed-effect floor at 0 lx, computed from the original data used to fit
  # the model, not from the new simulation design.
  model_frame <- stats::model.frame(model)

  if (!"lights_on" %in% names(model_frame)) {
    stop(
      "Could not compute baseline floor because `lights_on` is not in the model frame.",
      call. = FALSE
    )
  }

  model_frame <- model_frame %>%
    dplyr::mutate(
      lights_on = as.logical(lights_on)
    )

  baseline_model_frame <- model_frame %>%
    dplyr::filter(!lights_on)

  if (nrow(baseline_model_frame) == 0) {
    stop(
      "Could not compute baseline floor because the model frame contains no lights_on == FALSE rows.",
      call. = FALSE
    )
  }

  baseline_floor_log10 <- mean(
    predict(
      model,
      newdata = baseline_model_frame,
      re.form = NA,
      type = "link",
      allow.new.levels = TRUE
    ),
    na.rm = TRUE
  )

  if (!is.finite(baseline_floor_log10)) {
    stop(
      "Could not compute a finite baseline floor from the model frame.",
      call. = FALSE
    )
  }

  baseline_floor_raw <- 10^baseline_floor_log10

  dat <- dat %>%
    dplyr::mutate(
      # Expected response from fixed + random effects, before flooring
      expected_log10_unfloored = eta_fixed + eta_random,
  
      # Indicator: was this row affected by the floor?
      floored = expected_log10_unfloored < baseline_floor_log10,
  
      # Expected response after applying physiological floor
      expected_log10 = pmax(
        expected_log10_unfloored,
        baseline_floor_log10
      ),
  
      # Use the same residual-noise draw for both versions,
      # so flooring is the only difference
      residual_noise = stats::rnorm(
        n = dplyr::n(),
        mean = 0,
        sd = sd_resid
      ),
  
      simulated_log10_unfloored = expected_log10_unfloored + residual_noise,
      simulated_log10 = expected_log10 + residual_noise,
  
      simulated_emg_unfloored = 10^simulated_log10_unfloored,
      simulated_emg = 10^simulated_log10,
  
      baseline_floor_log10 = baseline_floor_log10,
      baseline_floor_emg = baseline_floor_raw
    )

  dat
}

# simulated_data <- simulate_from_design(
#   design_df = design_matrix,
#   model = m,
#   sd_scaling = 1,
#   baseline_pool = observed_log_baseline
# )
```

### Visualise simulations against original data

```{r visualise-simulations}
vis_sim_vs_original <- function(model,
                                driver = c("mel", "phot", "rho"),
                                sd_scaling = 1,
                                sim_seed = seed,
                                facet = FALSE) {
  
  driver <- match.arg(driver)
  
  x_log_var <- dplyr::case_when(
    driver == "mel" ~ "log_melanopic_edi",
    driver == "rho" ~ "log_rhodopic_edi",
    driver == "phot" ~ "log_illuminance"
  )
  
  x_raw_var <- dplyr::case_when(
    driver == "mel" ~ "melanopic_edi",
    driver == "rho" ~ "rhodopic_edi",
    driver == "phot" ~ "illuminance"
  )
  
  x_label <- dplyr::case_when(
    driver == "mel" ~ "Melanopic EDI",
    driver == "rho" ~ "Rhodopic EDI",
    driver == "phot" ~ "Illuminance"
  )
  
  model_frame <- model$frame
  
  baseline_floor_log10 <- mean(
    predict(
      model,
      newdata = model_frame %>%
        dplyr::mutate(lights_on = as.logical(lights_on)) %>%
        dplyr::filter(!lights_on),
      re.form = NA,
      type = "link",
      allow.new.levels = TRUE
    ),
    na.rm = TRUE
  )
  
  baseline_floor_emg <- 10^baseline_floor_log10
  
  original_df <- model_frame %>%
    dplyr::mutate(
      lights_on = as.logical(lights_on),
      x_plot = dplyr::if_else(
        lights_on,
        10^.data[[x_log_var]],
        0
      ),
      emg_response_plot = 10^log_emg_raw,
      source = "Original data",
      floored = FALSE
    ) %>%
    dplyr::select(
      x_plot,
      emg_response_plot,
      lights_on,
      floored,
      source
    )
  
  sim_df <- simulate_from_design(
    design_df = design_matrix,
    model = model,
    sd_scaling = sd_scaling,
    seed = sim_seed,
    baseline_pool = observed_log_baseline
  ) %>%
    dplyr::mutate(
      x_plot = .data[[x_raw_var]],
      lights_on = as.logical(lights_on)
    )
  
  sim_plot_df <- sim_df %>%
    tidyr::pivot_longer(
      cols = c(simulated_emg_unfloored, simulated_emg),
      names_to = "source",
      values_to = "emg_response_plot"
    ) %>%
    dplyr::mutate(
      source = dplyr::recode(
        source,
        simulated_emg_unfloored = "Simulated, no floor",
        simulated_emg = "Simulated, floor"
      )
    ) %>%
    dplyr::select(
      x_plot,
      emg_response_plot,
      lights_on,
      floored,
      source
    )
  
  plot_df <- dplyr::bind_rows(
    original_df,
    sim_plot_df
  ) %>%
    dplyr::mutate(
      source = factor(
        source,
        levels = c(
          "Original data",
          "Simulated, no floor",
          "Simulated, floor"
        )
      ),
      point_colour_group = dplyr::case_when(
        source == "Simulated, floor" & floored ~ "Simulated, floor — floored",
        source == "Simulated, floor" & !floored ~ "Simulated, floor — not floored",
        TRUE ~ as.character(source)
      ),
      smooth_group = point_colour_group
    )
  
  smooth_df <- plot_df %>%
    dplyr::filter(
      source != "Original data" | lights_on
    )
  
  baseline_label_df <- tibble::tibble(
    x_plot = 1,
    emg_response_plot = baseline_floor_emg,
    label = "floor",
    source = factor(
      "Original data",
      levels = c(
        "Original data",
        "Simulated, no floor",
        "Simulated, floor"
      )
    )
  )
  
  p <- ggplot2::ggplot(
    plot_df,
    ggplot2::aes(
      x = x_plot,
      y = emg_response_plot,
      colour = point_colour_group
    )
  ) +
    ggplot2::geom_point(alpha = 0.07) +
    ggplot2::geom_hline(
      yintercept = baseline_floor_emg,
      linetype = "dashed",
      linewidth = 0.8,
      colour = "black",
      alpha = 0.5
    ) +
    ggplot2::geom_text(
      data = baseline_label_df,
      ggplot2::aes(
        x = x_plot,
        y = emg_response_plot,
        label = label
      ),
      inherit.aes = FALSE,
      hjust = 1,
      vjust = -0.6,
      colour = "black",
      size = 3.5
    ) +
    ggplot2::geom_smooth(
      data = smooth_df,
      ggplot2::aes(
        group = smooth_group,
        colour = point_colour_group
      ),
      method = "lm",
      se = TRUE
    ) +
    ggplot2::scale_x_continuous(
      trans = LightLogR::symlog_trans(thr = 1),
      breaks = c(0, 1, 10, 100, 1000, 10000),
      minor_breaks = NULL
    ) +
    ggplot2::scale_y_log10(
      breaks = breaks,
      minor_breaks = minor_breaks
    ) +
    ggplot2::scale_colour_manual(
      values = c(
        "Original data" = "black",
        "Simulated, no floor" = "darkgreen",
        "Simulated, floor — not floored" = "orange",
        "Simulated, floor — floored" = "darkred"
      ),
      breaks = c(
        "Original data",
        "Simulated, no floor",
        "Simulated, floor — not floored",
        "Simulated, floor — floored"
      ),
      labels = c(
        "Original data",
        "Simulated, no floor",
        "Simulated, floor: unfloored data",
        "Simulated, floor: floored data"
      )
    ) +
    ggplot2::labs(
      x = x_label,
      y = "OO-EMG response",
      colour = NULL,
      title = glue::glue(
        "Original model-fitting data vs simulated data; driver = {driver}; SD scaling factor = {sd_scaling}"
      )
    ) +
    ggplot2::theme_minimal()
  
  if (facet) {
    p <- p +
      ggplot2::facet_wrap(~ source, scales = "fixed") +
      ggplot2::guides(colour = "none")
  }
  
  return(p)
}

vis_sim_vs_original(
  model = m_mel,
  driver = "mel",
  sd_scaling = 1, 
  facet = TRUE
)

vis_sim_vs_original(
  model = m_phot,
  driver = "phot",
  sd_scaling = 1, 
  facet = TRUE
)

vis_sim_vs_original(
  model = m_rho,
  driver = "rho",
  sd_scaling = 1, 
  facet = TRUE
)
```

To make it clear, for the power analysis, we will only use the yellow part
of the plots above, i.e. exclude the non-physiological data

This discarded data (red on the plots) reflects the stimulus design. Because the design is weighted toward the short-wavelength region, it favours melanopic and rhodopic stimulation more than photopic stimulation. Consequently, photopic illuminance reaches lower minimum values, down to approximately 2 lx, whereas melanopic and rhodopic EDI remain higher, at approximately 10 and 15 lx. Under a photopic generative model, these low illuminance values make simulated responses more likely to fall below the physiological lower bound. These responses are then discarded before model fitting. This explains why photopic-driver simulations show more discarded data, as illustrated in the final discard-summary plot.

# Collinearity of predictors

We first look at collinearity of melanopic EDI, illuminance and rhodopic EDI
If high collinearity, we cannot have all predictors in the same model…

```{r check-predictor-collinearity}
#| fig-width: 8
#| fig-height: 12

design_matrix %>%
  dplyr::filter(MEDI > 0, illuminance > 0) %>%
  dplyr::summarise(
    cor_raw = cor(illuminance, MEDI),
    cor_log = cor(log_illuminance, log_melanopic_edi),
    sd_log_difference = sd(log_illuminance - log_melanopic_edi)
  )

# Simulate data
sim_dat <- simulate_from_design(
  design_df = design_matrix, 
  model = m_mel, 
  sd_scaling = 1, 
  seed = seed, 
  baseline_pool = observed_log_baseline
)

formula_model_col <- log10(simulated_emg) ~ 
  log_illuminance + 
  log_melanopic_edi + 
  log_rhodopic_edi + 
  (1 | participant_id)

model_col <- lmer(formula_model_col, data = sim_dat)

summary(model_col)
performance::check_model(model_col)
cov2cor(vcov(model_col))
performance::check_collinearity(model_col)
```

Very high collinearity issues. Model struggles
The issue comes from melanopic and rhodopic curves being so close to each other.

We therefore cannot use such a model.

# Model fit comparisons

```{r model-fit-comparisons}
## Model fit comparisons with diagnostics
## glmmTMB only

fit_one_model <- function(formula, dat, model_name) {
  
  warnings <- character()
  error_msg <- NA_character_
  model <- NULL
  
  model <- tryCatch(
    withCallingHandlers(
      {
        glmmTMB::glmmTMB(
          formula = formula,
          data = dat,
          REML = FALSE
        )
      },
      warning = function(w) {
        warnings <<- c(warnings, conditionMessage(w))
        invokeRestart("muffleWarning")
      }
    ),
    error = function(e) {
      error_msg <<- conditionMessage(e)
      NULL
    }
  )
  
  if (is.null(model)) {
    return(
      list(
        model = NULL,
        diagnostics = tibble::tibble(
          model_name = model_name,
          converged = FALSE,
          singular = NA,
          convergence_code = NA_integer_,
          pdHess = NA,
          fit_message = NA_character_,
          warnings = paste(unique(warnings), collapse = " | "),
          error = error_msg,
          AIC = NA_real_
        )
      )
    )
  }
  
  convergence_code <- model$fit$convergence
  pdHess <- tryCatch(model$sdr$pdHess, error = function(e) NA)
  fit_message <- tryCatch(model$fit$message, error = function(e) NA_character_)
  
  converged <- isTRUE(convergence_code == 0) && isTRUE(pdHess)
  
  singular <- tryCatch(
    as.logical(performance::check_singularity(model)),
    error = function(e) NA
  )
  
  aic <- tryCatch(AIC(model), error = function(e) NA_real_)
  
  list(
    model = model,
    diagnostics = tibble::tibble(
      model_name = model_name,
      converged = converged,
      singular = singular,
      convergence_code = convergence_code,
      pdHess = pdHess,
      fit_message = fit_message,
      warnings = paste(unique(warnings), collapse = " | "),
      error = error_msg,
      AIC = aic
    )
  )
}

fit_models <- function(dat,
                       response = c("simulated_log10", "simulated_log10_unfloored")) {
  
  response <- match.arg(response)
  
  dat <- dat %>%
    dplyr::mutate(
      participant_id = factor(participant_id),
      trial = as.numeric(trial),
      lights_on = as.logical(lights_on)
    )
  
  mel_formula <- stats::as.formula(
    paste(
      response,
      "~ log_melanopic_edi +",
      "(1 | participant_id)"
    )
  )
  
  phot_formula <- stats::as.formula(
    paste(
      response,
      "~ log_illuminance +",
      "(1 | participant_id)"
    )
  )
  
  rho_formula <- stats::as.formula(
    paste(
      response,
      "~ log_rhodopic_edi +",
      "(1 | participant_id)"
    )
  )
  
  mel_fit <- fit_one_model(
    formula = mel_formula,
    dat = dat,
    model_name = "mel"
  )
  
  phot_fit <- fit_one_model(
    formula = phot_formula,
    dat = dat,
    model_name = "phot"
  )
  
  rho_fit <- fit_one_model(
    formula = rho_formula,
    dat = dat,
    model_name = "rho"
  )
  
  diagnostics <- dplyr::bind_rows(
    mel_fit$diagnostics,
    phot_fit$diagnostics,
    rho_fit$diagnostics
  )
  
  list(
    model_mel = mel_fit$model,
    model_phot = phot_fit$model,
    model_rho = rho_fit$model,
    diagnostics = diagnostics
  )
}

compute_delta_aic <- function(model_list, delta_aic_threshold = 2) {
  
  diag <- model_list$diagnostics
  
  required_models <- c("mel", "phot", "rho")
  
  model_diag <- diag %>%
    dplyr::filter(model_name %in% required_models)
  
  all_converged <- model_diag %>%
    dplyr::summarise(
      ok = all(converged[match(required_models, model_name)])
    ) %>%
    dplyr::pull(ok)
  
  if (!isTRUE(all_converged)) {
    return(
      tibble::tibble(
        AIC_mel = NA_real_,
        AIC_phot = NA_real_,
        AIC_rho = NA_real_,
        best_model = NA_character_,
        delta_AIC_mel = NA_real_,
        delta_AIC_phot = NA_real_,
        delta_AIC_rho = NA_real_,
        second_best_delta_AIC = NA_real_,
        evidence = NA_character_
      )
    )
  }
  
  aic_summary <- model_diag %>%
    dplyr::select(model_name, AIC) %>%
    tidyr::pivot_wider(
      names_from = model_name,
      values_from = AIC,
      names_prefix = "AIC_"
    ) %>%
    dplyr::rowwise() %>%
    dplyr::mutate(
      min_AIC = min(c(AIC_mel, AIC_phot, AIC_rho), na.rm = TRUE),
      
      delta_AIC_mel = AIC_mel - min_AIC,
      delta_AIC_phot = AIC_phot - min_AIC,
      delta_AIC_rho = AIC_rho - min_AIC,
      
      best_model = c("mel", "phot", "rho")[
        which.min(c(AIC_mel, AIC_phot, AIC_rho))
      ],
      
      second_best_delta_AIC = sort(c(
        delta_AIC_mel,
        delta_AIC_phot,
        delta_AIC_rho
      ))[2],
      
      evidence = dplyr::case_when(
        second_best_delta_AIC > delta_aic_threshold ~
          paste0("Favours ", best_model, " model"),
        TRUE ~
          "No clear preference"
      )
    ) %>%
    dplyr::ungroup() %>%
    dplyr::select(
      AIC_mel,
      AIC_phot,
      AIC_rho,
      best_model,
      delta_AIC_mel,
      delta_AIC_phot,
      delta_AIC_rho,
      second_best_delta_AIC,
      evidence
    )
  
  aic_summary
}
```

Small note:
We dropped random slopes here because they were causing convergence issues.

# Run loop

The analysis uses up to 6 cores, depending on the CPU the code is being run on

```{r run-loop, warning=FALSE}
n_simulations <- 1000

available_cores <- parallel::detectCores(logical = FALSE)

# As the line above can return NA:
if (is.na(available_cores)) {
  available_cores <- parallel::detectCores(logical = TRUE)
}

if (is.na(available_cores)) {
  available_cores <- 2
}

# Never use more than 6 cores
n_workers <- min(
  6,
  max(1, available_cores - 1)
)

message("Using ", n_workers, " parallel workers.")

future::plan(
  future::multisession,
  workers = n_workers
)

message("Using ", n_workers, " parallel workers.")

future::plan(
  future::multisession,
  workers = n_workers
)

options(progressr.enable = TRUE)

run_aic_sim <- function(
    n_sims = 1000,
    n_subjects = 20,
    noise_scaling = 1,
    driver,
    gen_model,
    delta_aic_threshold = 2
) {
  
  response <- "simulated_log10"
  
  available_subjects <- design_matrix %>%
    dplyr::distinct(participant_id) %>%
    dplyr::arrange(participant_id) %>%
    dplyr::pull(participant_id)
  
  if (n_subjects > length(available_subjects)) {
    stop(
      "Requested ", n_subjects, " participants, but design_matrix only contains ",
      length(available_subjects), ".",
      call. = FALSE
    )
  }
  
  selected_subjects <- sample(available_subjects, n_subjects, replace = FALSE)
  
  design_subset <- design_matrix %>%
    dplyr::filter(participant_id %in% selected_subjects) %>%
    dplyr::mutate(
      participant_id = factor(participant_id)
    )
  
  sim_results <- purrr::map(
    seq_len(n_sims),
    ~ {
      sim_data <- simulate_from_design(
        design_df = design_subset,
        model = gen_model,
        sd_scaling = noise_scaling,
        seed = NULL,
        baseline_pool = observed_log_baseline
      )
      
      # Track how much data would be discarded because it hit the physiological floor
      n_rows_before_discard <- nrow(sim_data)
      n_rows_floored <- sum(sim_data$floored, na.rm = TRUE)
      prop_floored <- mean(sim_data$floored, na.rm = TRUE)
      
      sim_data <- sim_data %>% 
        dplyr::filter(floored == FALSE)
      
      n_rows_after_discard <- nrow(sim_data)
      
      fits <- fit_models(
        dat = sim_data,
        response = response
      )
      
      aic_result <- compute_delta_aic(
        fits,
        delta_aic_threshold = delta_aic_threshold
      )
      
      diagnostics <- fits$diagnostics %>%
        dplyr::mutate(
          sim_id = .x,
          driver = driver,
          n_subjects = n_subjects,
          noise_scaling = noise_scaling,
          response = response
        )
      
      list(
        aic_result = aic_result %>%
          dplyr::mutate(
            sim_id = .x,
            driver = driver,
            n_subjects = n_subjects,
            noise_scaling = noise_scaling,
            response = response,
            n_rows_before_discard = n_rows_before_discard,
            n_rows_floored = n_rows_floored,
            n_rows_after_discard = n_rows_after_discard,
            prop_floored = prop_floored
          ),
        diagnostics = diagnostics
      )
    }
  )
  
  aic_results <- purrr::map_dfr(sim_results, "aic_result")
  
  diagnostics_summary <- purrr::map_dfr(sim_results, "diagnostics") %>%
    dplyr::group_by(driver, n_subjects, noise_scaling, response, model_name) %>%
    dplyr::summarise(
      prop_converged = mean(converged, na.rm = TRUE),
      prop_singular = mean(singular, na.rm = TRUE),
      n_fits = dplyr::n(),
      .groups = "drop"
    )
  
  evidence_list <- aic_results$evidence
  
  n_valid <- sum(!is.na(evidence_list))
  n_failed <- sum(is.na(evidence_list))
  
  if (n_valid == 0) {
    summary <- tibble::tibble(
      p_favour_mel = NA_real_,
      p_favour_phot = NA_real_,
      p_favour_rho = NA_real_,
      p_unclear = NA_real_,
      n_sims = n_sims,
      n_valid = n_valid,
      n_failed = n_failed,
      prop_failed = n_failed / n_sims,
      driver = driver,
      n_subjects = n_subjects,
      noise_scaling = noise_scaling,
      response = response
    )
  } else {
    summary <- tibble::tibble(
      p_favour_mel = mean(evidence_list == "Favours mel model", na.rm = TRUE),
      p_favour_phot = mean(evidence_list == "Favours phot model", na.rm = TRUE),
      p_favour_rho = mean(evidence_list == "Favours rho model", na.rm = TRUE),
      p_unclear = mean(evidence_list == "No clear preference", na.rm = TRUE),
      n_sims = n_sims,
      n_valid = n_valid,
      n_failed = n_failed,
      prop_failed = n_failed / n_sims,
      driver = driver,
      n_subjects = n_subjects,
      noise_scaling = noise_scaling,
      response = response
    )
  }
  
  list(
    summary = summary,
    aic_results = aic_results,
    diagnostics_summary = diagnostics_summary
  )
}

grid <- tidyr::expand_grid(
  driver_models,
  n_subjects = as.integer(seq(2, 50)), # 2 to 50 participants
  noise_scaling = c(0.5, 1, 2, 4, 6, 8, 10)
)

aic_nested <- progressr::with_progress(
  {
    p <- progressr::progressor(steps = nrow(grid))
    
    results <- furrr::future_map(
      seq_len(nrow(grid)),
      function(i) {
        out <- run_aic_sim(
          n_sims = n_simulations,
          driver = grid$driver[[i]],
          gen_model = grid$model[[i]],
          n_subjects = grid$n_subjects[[i]],
          noise_scaling = grid$noise_scaling[[i]]
        )
        
        p(sprintf(
          "completed: driver = %s | subjects = %s | noise = %s",
          grid$driver[[i]],
          grid$n_subjects[[i]],
          grid$noise_scaling[[i]]
        ))
        
        out
      },
      .options = furrr::furrr_options(
        seed = TRUE,
        scheduling = Inf,
        stdout = TRUE,
        conditions = "condition"
      )
    )
    
    grid %>%
      dplyr::mutate(results = results)
  },
  handlers = progressr::handler_txtprogressbar()
)

aic_diagnostics_summary <- aic_nested %>%
  dplyr::transmute(
    diagnostics_summary = purrr::map(results, "diagnostics_summary")
  ) %>%
  tidyr::unnest(diagnostics_summary)

aic_per_sim <- aic_nested %>%
  dplyr::transmute(
    driver,
    n_subjects,
    noise_scaling,
    aic_results = purrr::map(
      results,
      ~ .x$aic_results %>%
        dplyr::select(
          -dplyr::any_of(c("driver", "n_subjects", "noise_scaling"))
        )
    )
  ) %>%
  tidyr::unnest(aic_results)

discard_summary <- aic_per_sim %>%
  dplyr::group_by(driver, n_subjects, noise_scaling, response) %>%
  dplyr::summarise(
    mean_prop_floored = mean(prop_floored, na.rm = TRUE),
    sd_prop_floored = sd(prop_floored, na.rm = TRUE),
    median_prop_floored = median(prop_floored, na.rm = TRUE),
    min_prop_floored = min(prop_floored, na.rm = TRUE),
    max_prop_floored = max(prop_floored, na.rm = TRUE),
    mean_n_rows_before_discard = mean(n_rows_before_discard, na.rm = TRUE),
    mean_n_rows_floored = mean(n_rows_floored, na.rm = TRUE),
    mean_n_rows_after_discard = mean(n_rows_after_discard, na.rm = TRUE),
    .groups = "drop"
  )

discard_by_driver <- discard_summary %>%
  dplyr::group_by(driver) %>%
  dplyr::summarise(
    mean_discarded_pct = mean(mean_prop_floored, na.rm = TRUE) * 100,
    sd_discarded_pct = sd(mean_prop_floored, na.rm = TRUE) * 100,
    min_discarded_pct = min(mean_prop_floored, na.rm = TRUE) * 100,
    max_discarded_pct = max(mean_prop_floored, na.rm = TRUE) * 100,
    .groups = "drop"
  )

evidence_levels <- c(
  "Favours mel model",
  "Favours phot model",
  "Favours rho model",
  "No clear preference"
)

evidence_props <- aic_per_sim %>%
  dplyr::filter(!is.na(evidence)) %>%
  dplyr::mutate(
    evidence = factor(evidence, levels = evidence_levels)
  ) %>%
  dplyr::count(
    driver,
    noise_scaling,
    n_subjects,
    evidence,
    name = "n",
    .drop = FALSE
  ) %>%
  dplyr::group_by(driver, noise_scaling, n_subjects) %>%
  tidyr::complete(
    evidence = factor(evidence_levels, levels = evidence_levels),
    fill = list(n = 0)
  ) %>%
  dplyr::mutate(
    prop = n / sum(n)
  ) %>%
  dplyr::ungroup()

glimpse(aic_diagnostics_summary)
glimpse(aic_per_sim)
glimpse(discard_summary)
glimpse(discard_by_driver)
glimpse(evidence_props)
```

```{r save-results}
output_dir <- here::here("outputs", "power_analysis")

dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)

saveRDS(
  aic_per_sim,
  file.path(output_dir, "power_analysis_full_results.rds")
)

readr::write_csv(
  aic_diagnostics_summary,
  file.path(output_dir, "power_analysis_diagnostics_summary.csv")
)

readr::write_csv(
  discard_summary,
  file.path(output_dir, "power_analysis_discard_summary.csv")
)

readr::write_csv(
  evidence_props,
  file.path(output_dir, "power_analysis_evidence_props.csv")
)
```

Proportion of discarded simulated data does not depend on noise scaling or
sample size:

```{r}
# discard_summary_clean <- discard_summary %>%
#   filter(response == "simulated_log10") %>%
#   mutate(
#     n_subjects_c = scale(n_subjects, center = TRUE, scale = TRUE)[, 1],
#     noise_scaling_c = scale(noise_scaling, center = TRUE, scale = TRUE)[, 1]
#   )

m_lm <- lm(
  mean_prop_floored ~ driver + n_subjects + noise_scaling,
  data = discard_summary
)

summary(m_lm)
confint(m_lm)
tidy(m_lm, conf.int = TRUE)
```


# Plotting results

```{r reload-data}
# aic_per_sim <- readRDS(here::here("outputs", "power_analysis", "power_analysis_full_results.csv"))

# aic_diagnostics_summary <- read.csv(here::here("outputs", "power_analysis", "power_analysis_diagnostics_summary.csv"))
# discard_summary <- read.csv(here::here("outputs", "power_analysis", "power_analysis_discard_summary.csv"))
# evidence_props <- read.csv(here::here("outputs", "power_analysis", "power_analysis_evidence_props.csv"))
```


```{r plot-results}
chosen_n <- 20
chosen_n_position <- which(sort(unique(evidence_props$n_subjects)) == chosen_n)

evidence_colours <- c(
  "Favours mel model" = "#009E73",      # bluish green
  "Favours phot model" = "#D55E00",     # vermillion
  "Favours rho model" = "#0072B2",      # blue
  "No clear preference" = "grey60"
)

model_colours <- c(
  "mel" = "#009E73",   # bluish green
  "phot" = "#D55E00",  # vermillion
  "rho" = "#0072B2"    # blue
)

clean_facet_labels <- ggplot2::labeller(
  driver = c(
    mel = "Driver: melanopic EDI",
    phot = "Driver: illuminance",
    rho = "Driver: rhodopic EDI"
  ),
  noise_scaling = function(x) paste0("Noise scaling: ", x)
)

plot_power <- evidence_props %>%
  ggplot2::ggplot(
    ggplot2::aes(
      x = n_subjects,
      y = prop,
      colour = evidence,
      group = evidence
    )
  ) +
  ggplot2::geom_vline(
    ggplot2::aes(
      xintercept = chosen_n,
      linetype = "Chosen sample size"
    ),
    colour = "grey30",
    linewidth = 0.7
  ) +
  ggplot2::scale_linetype_manual(
    name = NULL,
    values = c("Chosen sample size" = "dashed")
  ) +
  ggplot2::geom_point(size = 2) +
  ggplot2::geom_line(linewidth = 0.8) +
  ggplot2::facet_grid(
    driver ~ noise_scaling,
    labeller = clean_facet_labels
  ) +
  ggplot2::scale_y_continuous(
    limits = c(0, 1),
    breaks = seq(0, 1, by = 0.25)
  ) +
  ggplot2::scale_colour_manual(
    values = evidence_colours,
    drop = FALSE
  ) +
  ggplot2::labs(
    x = "Number of participants",
    y = "Proportion of simulations",
    colour = "Evidence",
    title = "Evidence outcome proportions by assumed generative driver"
  ) +
  ggplot2::theme_minimal() +
  plot_formatting

plot_power_2 <- evidence_props %>%
  ggplot2::ggplot(
    ggplot2::aes(
      x = factor(n_subjects),
      y = prop,
      fill = evidence
    )
  ) +
  ggplot2::geom_vline(
    ggplot2::aes(
      xintercept = chosen_n_position,
      linetype = "Chosen sample size"
    ),
    colour = "grey30",
    linewidth = 0.7
  ) +
  ggplot2::scale_linetype_manual(
    name = NULL,
    values = c("Chosen sample size" = "dashed")
  ) +
  ggplot2::geom_col(width = 0.8) +
  ggplot2::facet_grid(
    driver ~ noise_scaling,
    labeller = clean_facet_labels
  ) +
  ggplot2::scale_y_continuous(
    limits = c(0, 1),
    breaks = seq(0, 1, by = 0.25)
  ) +
  ggplot2::scale_fill_manual(
    values = evidence_colours,
    drop = FALSE
  ) +
  ggplot2::labs(
    x = "Number of participants",
    y = "Proportion of simulations",
    fill = "Evidence",
    title = "Evidence outcome proportions by assumed generative driver"
  ) +
  ggplot2::theme_minimal() +
  plot_formatting

plot_convergence <- aic_diagnostics_summary %>%
  ggplot2::ggplot(
    ggplot2::aes(
      x = n_subjects,
      y = prop_converged,
      colour = model_name,
      group = model_name
    )
  ) +
  ggplot2::geom_vline(
    ggplot2::aes(
      xintercept = chosen_n,
      linetype = "Chosen sample size"
    ),
    colour = "grey30",
    linewidth = 0.7
  ) +
  ggplot2::geom_point(size = 2, alpha = 0.3) +
  ggplot2::geom_line(linewidth = 0.8) +
  ggplot2::facet_grid(
    driver ~ noise_scaling,
    labeller = clean_facet_labels
  ) +
  ggplot2::scale_colour_manual(
    values = model_colours,
    drop = FALSE
  ) +
  ggplot2::scale_linetype_manual(
    name = NULL,
    values = c("Chosen sample size" = "dashed")
  ) +
  ggplot2::scale_y_continuous(
    limits = c(0, 1),
    breaks = seq(0, 1, by = 0.25)
  ) +
  ggplot2::labs(
    x = "Number of participants",
    y = "Proportion of converged fits",
    colour = "Fitted model",
    title = "Convergence rate by driver, sample size, and noise level"
  ) +
  ggplot2::theme_minimal() +
  plot_formatting

plot_discarded <- discard_summary %>%
  ggplot2::ggplot(
    ggplot2::aes(
      x = n_subjects,
      y = factor(noise_scaling),
      fill = mean_prop_floored
    )
  ) +
  ggplot2::geom_tile() +
  ggplot2::geom_vline(
    ggplot2::aes(
      xintercept = chosen_n,
      linetype = "Chosen sample size"
    ),
    colour = "grey30",
    linewidth = 0.7
  ) +
  ggplot2::scale_linetype_manual(
    name = NULL,
    values = c("Chosen sample size" = "dashed")
  ) +
  ggplot2::scale_fill_viridis_c(
    labels = scales::percent_format(accuracy = 1),
    limits = c(0, NA)
  ) +
  ggplot2::facet_wrap(
    ~ driver,
    labeller = clean_facet_labels
  ) +
  ggplot2::labs(
    x = "Number of participants",
    y = "Noise scaling",
    fill = "Discarded rows",
    title = "Proportion of simulated rows discarded as non-physiological"
  ) +
  ggplot2::theme_minimal() +
  plot_formatting

plot_power
plot_power_2
plot_convergence
plot_discarded
```


```{r save-plots}
save_rds_safe(
  plot_power,
  here::here("outputs", "figures", "rds", glue::glue("power_analysis_evidence.rds"))
)

formatted_save(
  here::here("outputs", "figures", "power_analysis_evidence.jpg"), 
  plot_power, 
  width = 9, 
  height = 7
)

save_rds_safe(
  plot_power_2,
  here::here("outputs", "figures", "rds", glue::glue("power_analysis_evidence_bar.rds"))
)

formatted_save(
  here::here("outputs", "figures", "power_analysis_evidence_bar.jpg"), 
  plot_power_2, 
  width = 9, 
  height = 7
)

save_rds_safe(
  plot_convergence,
  here::here("outputs", "figures", "rds", glue::glue("power_analysis_convergence.rds"))
)

formatted_save(
  here::here("outputs", "figures", "power_analysis_convergence.jpg"), 
  plot_convergence, 
  width = 9, 
  height = 7
)

save_rds_safe(
  plot_discarded,
  here::here(
    "outputs",
    "figures",
    "rds",
    glue::glue("power_analysis_discarded_data.rds")
  )
)

formatted_save(
  here::here("outputs", "figures", "power_analysis_discarded_data.jpg"), 
  plot_discarded, 
  width = 9, 
  height = 7
)
```

```{r session-info}
sessionInfo()
```

```{r render-time-2, echo=FALSE}
render_time <- Sys.time() - .render_start
cat("Render time:", format(render_time))
```
