---
title: "BFDA from pilot data model fit"
author: "Lucien Bickerstaff"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r render-time-1, include=FALSE}
# Record how long it took to render the file
.render_start <- Sys.time()
```

```{r setup, warning=F, message=F}
# Seed for reproducibility
set.seed(2025)

# Libraries
library(here)
library(glue)
library(broom.mixed)
library(jsonlite)
library(BayesFactor)
library(lme4)
library(MASS)
library(tidyverse)
library(furrr)
library(patchwork)
library(ggtext)

# Paths
source(here("R", "helpers.R"))
source(here("R", "plot_settings.R"))
```

# Load and prepare data

## Load models

We use EMG as our reference modality, as it is the most reliable and the most indicative

```{r}
m <- readRDS(here("outputs", "emg_lmm.rds"))
```

## Load design matrix

from compute_primary_settings.Rmd

```{r}
design_matrix <- readr::read_csv(
  here::here("outputs", "design_matrix.csv"),
  show_col_types = FALSE
) %>%
  dplyr::mutate(
    participant_id = factor(participant_id),
    lab_visit = factor(lab_visit),

    melanopic_edi = MEDI,

    # Must match the transformation used when fitting the EMG model.
    # This keeps the 0-lx condition while avoiding log10(0).
    log_illuminance = log10(pmax(illuminance, 1)),
    log_melanopic_edi = log10(pmax(melanopic_edi, 1)),

    lights_on = melanopic_edi != 0,
  )
```


```{r}
# Simulation matrix also needs `log_baseline`

# -----------------------------
# Simulate log_baseline using a shifted gamma distribution
# -----------------------------

model_frame <- model.frame(m)

observed_log_baseline <- model_frame$log_baseline
observed_log_baseline <- observed_log_baseline[is.finite(observed_log_baseline)]

# Shift log_baseline so all values are positive
shift <- abs(min(observed_log_baseline)) + 0.001

log_baseline_shifted <- observed_log_baseline + shift

# Fit gamma parameters on shifted log_baseline
baseline_mean <- mean(log_baseline_shifted)
baseline_var <- var(log_baseline_shifted)

gamma_shape <- baseline_mean^2 / baseline_var
gamma_rate <- baseline_mean / baseline_var

if (!is.finite(gamma_shape) || !is.finite(gamma_rate) ||
    gamma_shape <= 0 || gamma_rate <= 0) {
  stop("Invalid shifted-gamma parameters for log_baseline.", call. = FALSE)
}

set.seed(123)

design_matrix <- design_matrix %>%
  dplyr::mutate(
    log_baseline = rgamma(
      n = dplyr::n(),
      shape = gamma_shape,
      rate = gamma_rate
    ) - shift
  )

# Observed vs simulated
observed_plot_df <- tibble::tibble(
  source = "Observed",
  log_baseline = observed_log_baseline
)

simulated_plot_df <- design_matrix %>%
  dplyr::select(log_baseline) %>%
  dplyr::mutate(source = "Simulated")

plot_df <- dplyr::bind_rows(observed_plot_df, simulated_plot_df)

ggplot(plot_df, aes(x = log_baseline)) +
  geom_histogram(bins = 30, colour = "white") +
  facet_wrap(~ source, scales = "free_y") +
  theme_classic() +
  labs(
    x = "log_baseline",
    y = "Count",
    title = "Observed vs simulated log_baseline"
  )
```

# Define functions

## Data simulation

```{r}
simulate_from_design <- function(design_df,
                                 model,
                                 sd_scaling = 1,
                                 seed = NULL) {
  if (!is.null(seed)) {
    set.seed(seed)
  }

  if (!requireNamespace("MASS", quietly = TRUE)) {
    stop("Package 'MASS' is required.", call. = FALSE)
  }

  if (!requireNamespace("lme4", quietly = TRUE)) {
    stop("Package 'lme4' is required.", call. = FALSE)
  }

  if (!requireNamespace("glmmTMB", quietly = TRUE)) {
    stop("Package 'glmmTMB' is required.", call. = FALSE)
  }

  # ---------------------------------------------------------------------------
  # Prepare and check design matrix
  # ---------------------------------------------------------------------------

  dat <- design_df %>%
    dplyr::mutate(
      participant_id = factor(participant_id),
      trial = as.numeric(trial),
      lights_on = as.logical(lights_on)
    )

  # Variables needed by the model RHS, including random-effect grouping variables
  required_vars <- all.vars(stats::formula(model)[[3]])

  missing_vars <- setdiff(required_vars, names(dat))

  if (length(missing_vars) > 0) {
    stop(
      "The design matrix is missing variables required by the model: ",
      paste(missing_vars, collapse = ", "),
      call. = FALSE
    )
  }

  # Check for missing predictor values
  predictor_na_counts <- dat %>%
    dplyr::select(dplyr::all_of(required_vars)) %>%
    dplyr::summarise(
      dplyr::across(
        dplyr::everything(),
        ~ sum(is.na(.x))
      )
    )

  if (any(predictor_na_counts > 0)) {
    print(predictor_na_counts)
    stop(
      "Some model predictors contain NA values in the design matrix.",
      call. = FALSE
    )
  }

  # ---------------------------------------------------------------------------
  # Fixed-effect prediction
  # ---------------------------------------------------------------------------

  # Fixed-effects-only prediction on the model scale.
  # For a Gaussian identity-link model fitted to log_emg_raw, this is the
  # predicted log_emg_raw value without participant-specific random effects.
  dat$eta_fixed <- predict(
    model,
    newdata = dat,
    re.form = NA,
    type = "link",
    allow.new.levels = TRUE
  )

  # ---------------------------------------------------------------------------
  # Simulate random effects
  # ---------------------------------------------------------------------------

  dat$eta_random <- 0

  random_terms <- lme4::findbars(stats::formula(model))

  vc <- glmmTMB::VarCorr(model)$cond

  for (term in random_terms) {

    random_effect_formula <- term[[2]]
    grouping_factor_expr  <- term[[3]]

    grouping_vars <- all.vars(grouping_factor_expr)
    grouping_name <- paste(grouping_vars, collapse = ":")

    if (!grouping_name %in% names(vc)) {
      warning(
        "Random-effect grouping factor '", grouping_name,
        "' not found in VarCorr(model). Skipping this random effect."
      )
      next
    }

    vc_matrix <- as.matrix(vc[[grouping_name]])

    # Examples:
    #   (1 | participant_id) becomes ~ 1
    #   (log_melanopic_edi | participant_id) becomes ~ log_melanopic_edi
    # Because model.matrix(~ x) includes the intercept, this correctly gives
    # random intercept + random slope for (x | participant_id).
    re_formula <- stats::as.formula(
      paste("~", paste(deparse(random_effect_formula), collapse = ""))
    )

    Z <- stats::model.matrix(re_formula, data = dat)

    group_id <- if (length(grouping_vars) == 1) {
      dat[[grouping_vars]]
    } else {
      interaction(dat[grouping_vars], drop = TRUE)
    }

    group_id <- factor(group_id)

    simulated_re <- MASS::mvrnorm(
      n = nlevels(group_id),
      mu = rep(0, ncol(vc_matrix)),
      Sigma = vc_matrix
    )

    simulated_re <- as.matrix(simulated_re)

    colnames(simulated_re) <- colnames(vc_matrix)
    rownames(simulated_re) <- levels(group_id)

    common_cols <- intersect(colnames(Z), colnames(simulated_re))

    if (length(common_cols) == 0) {
      warning(
        "No matching random-effect columns for grouping factor '",
        grouping_name,
        "'. Skipping this random effect."
      )
      next
    }

    re_contribution <- rowSums(
      Z[, common_cols, drop = FALSE] *
        simulated_re[as.character(group_id), common_cols, drop = FALSE]
    )

    dat$eta_random <- dat$eta_random + re_contribution
  }

  # ---------------------------------------------------------------------------
  # Simulate residual noise and back-transform response
  # Floor is applied to the expected response before residual noise is added
  # ---------------------------------------------------------------------------

  sd_resid <- stats::sigma(model) * sd_scaling

  # Fixed-effect floor at 0 lx, computed from the original data used to fit
  # the model, not from the new simulation design.
  model_frame <- stats::model.frame(model)

  if (!"lights_on" %in% names(model_frame)) {
    stop(
      "Could not compute baseline floor because `lights_on` is not in the model frame.",
      call. = FALSE
    )
  }

  model_frame <- model_frame %>%
    dplyr::mutate(
      lights_on = as.logical(lights_on)
    )

  baseline_model_frame <- model_frame %>%
    dplyr::filter(!lights_on)

  if (nrow(baseline_model_frame) == 0) {
    stop(
      "Could not compute baseline floor because the model frame contains no lights_on == FALSE rows.",
      call. = FALSE
    )
  }

  baseline_floor_log10 <- mean(
    predict(
      model,
      newdata = baseline_model_frame,
      re.form = NA,
      type = "link",
      allow.new.levels = TRUE
    ),
    na.rm = TRUE
  )

  if (!is.finite(baseline_floor_log10)) {
    stop(
      "Could not compute a finite baseline floor from the model frame.",
      call. = FALSE
    )
  }

  baseline_floor_raw <- 10^baseline_floor_log10

  dat <- dat %>%
    dplyr::mutate(
      # Expected response from fixed + random effects, before flooring
      expected_log10_unfloored = eta_fixed + eta_random,
  
      # Indicator: was this row affected by the floor?
      floored = expected_log10_unfloored < baseline_floor_log10,
  
      # Expected response after applying physiological floor
      expected_log10 = pmax(
        expected_log10_unfloored,
        baseline_floor_log10
      ),
  
      # Use the same residual-noise draw for both versions,
      # so flooring is the only difference
      residual_noise = stats::rnorm(
        n = dplyr::n(),
        mean = 0,
        sd = sd_resid
      ),
  
      simulated_log10_unfloored = expected_log10_unfloored + residual_noise,
      simulated_log10 = expected_log10 + residual_noise,
  
      simulated_emg_unfloored = 10^simulated_log10_unfloored,
      simulated_emg = 10^simulated_log10,
  
      baseline_floor_log10 = baseline_floor_log10,
      baseline_floor_emg = baseline_floor_raw
    )

  dat
}

simulated_data <- simulate_from_design(
  design_df = design_matrix,
  model = m,
  sd_scaling = 1,
  seed = 123
)
```


```{r}
vis_sim <- function(sd_scaling = 1,
                    seed = 123,
                    floored = TRUE,
                    compare_floor = FALSE) {
  
  sim_df <- simulate_from_design(
    design_df = design_matrix,
    model = m,
    sd_scaling = sd_scaling,
    seed = seed
  ) %>%
    dplyr::mutate(
      melanopic_edi_plot = pmax(melanopic_edi, 1)
    )
  
  if (compare_floor) {
    
    plot_df <- sim_df %>%
      tidyr::pivot_longer(
        cols = c(simulated_emg_unfloored, simulated_emg),
        names_to = "flooring",
        values_to = "emg_response_plot"
      ) %>%
      dplyr::mutate(
        flooring = dplyr::recode(
          flooring,
          simulated_emg_unfloored = "Unfloored + residual noise",
          simulated_emg = "Floored + residual noise"
        )
      )
    
    plot_title <- glue::glue(
      "Simulated OO-EMG response; SD scaling factor = {sd_scaling}"
    )
    
    y_label <- "Simulated OO-EMG response"
    
  } else {
    
    if (floored) {
      plot_df <- sim_df %>%
        dplyr::mutate(
          flooring = "Floored + residual noise",
          emg_response_plot = simulated_emg
        )
    } else {
      plot_df <- sim_df %>%
        dplyr::mutate(
          flooring = "Unfloored + residual noise",
          emg_response_plot = simulated_emg_unfloored
        )
    }
    
    plot_title <- glue::glue(
      "{unique(plot_df$flooring)}; SD scaling factor = {sd_scaling}"
    )
    
    y_label <- "Simulated OO-EMG response"
  }
  
  p <- plot_df %>%
    ggplot2::ggplot(
      ggplot2::aes(
        x = melanopic_edi_plot,
        y = emg_response_plot,
        colour = participant_id
      )
    ) +
    ggplot2::geom_point(alpha = 0.5) +
    ggplot2::geom_smooth(
      ggplot2::aes(colour = NULL),
      method = "lm",
      se = TRUE,
      color = "black"
    ) +
    ggplot2::scale_x_log10(breaks = breaks, minor_breaks = minor_breaks) +
    ggplot2::scale_y_log10(breaks = breaks, minor_breaks = minor_breaks) +
    ggplot2::labs(
      x = "Melanopic EDI, lux",
      y = y_label,
      colour = "Participant",
      title = plot_title
    ) +
    ggplot2::theme_minimal()
  
  if (compare_floor) {
    p <- p +
      ggplot2::facet_wrap(~ flooring)
  }
  
  print(p)
}

# changing variance in the data
vis_sim(sd_scaling = 0.1, floored = TRUE)
vis_sim(sd_scaling = 0.5, floored = TRUE)
vis_sim(sd_scaling = 1, floored = TRUE)
vis_sim(sd_scaling = 5, floored = TRUE)

vis_sim(sd_scaling = 1, floored = FALSE)

vis_sim(sd_scaling = 1, compare_floor = TRUE)
```

```{r}
vis_sim_vs_original <- function(sd_scaling = 1,
                                seed = 123,
                                facet = FALSE) {
  
  model_frame <- model.frame(m)
  
  baseline_floor_log10 <- mean(
    predict(
      m,
      newdata = model_frame %>%
        dplyr::mutate(lights_on = as.logical(lights_on)) %>%
        dplyr::filter(!lights_on),
      re.form = NA,
      type = "link",
      allow.new.levels = TRUE
    ),
    na.rm = TRUE
  )
  
  baseline_floor_emg <- 10^baseline_floor_log10
  
  original_df <- model_frame %>%
    dplyr::mutate(
      lights_on = as.logical(lights_on),
      
      melanopic_edi_plot = dplyr::if_else(
        lights_on,
        10^log_melanopic_edi,
        0
      ),
      
      emg_response_plot = 10^log_emg_raw,
      source = "Original data",
      floored = FALSE
    ) %>%
    dplyr::select(
      melanopic_edi_plot,
      emg_response_plot,
      lights_on,
      floored,
      source
    )
  
  sim_df <- simulate_from_design(
    design_df = design_matrix,
    model = m,
    sd_scaling = sd_scaling,
    seed = seed
  ) %>%
    dplyr::mutate(
      melanopic_edi_plot = melanopic_edi,
      lights_on = as.logical(lights_on)
    )
  
  sim_plot_df <- sim_df %>%
    tidyr::pivot_longer(
      cols = c(simulated_emg_unfloored, simulated_emg),
      names_to = "source",
      values_to = "emg_response_plot"
    ) %>%
    dplyr::mutate(
      source = dplyr::recode(
        source,
        simulated_emg_unfloored = "Simulated, no floor",
        simulated_emg = "Simulated, floor"
      )
    ) %>%
    dplyr::select(
      melanopic_edi_plot,
      emg_response_plot,
      lights_on,
      floored,
      source
    )
  
  plot_df <- dplyr::bind_rows(
    original_df,
    sim_plot_df
  ) %>%
    dplyr::mutate(
      source = factor(
        source,
        levels = c(
          "Original data",
          "Simulated, no floor",
          "Simulated, floor"
        )
      )
    ) %>%
    dplyr::mutate(
      smooth_group = dplyr::case_when(
        source == "Simulated, floor" & floored ~ "Simulated, floor: floored",
        source == "Simulated, floor" & !floored ~ "Simulated, floor: not floored",
        TRUE ~ as.character(source)
      )
    ) %>% 
    dplyr::mutate(
      point_colour_group = dplyr::case_when(
        source == "Simulated, floor" & floored ~ "Simulated, floor — floored",
        source == "Simulated, floor" & !floored ~ "Simulated, floor — not floored",
        TRUE ~ as.character(source)
      ),
      smooth_group = dplyr::case_when(
        source == "Simulated, floor" & floored ~ "Simulated, floor — floored",
        source == "Simulated, floor" & !floored ~ "Simulated, floor — not floored",
        TRUE ~ as.character(source)
      )
    )
  
  smooth_df <- plot_df %>%
    dplyr::filter(
      source != "Original data" | lights_on
    )
  
  baseline_label_df <- tibble::tibble(
    melanopic_edi_plot = 1,
    emg_response_plot = baseline_floor_emg,
    label = "floor",
    source = factor(
      "Original data",
      levels = c(
        "Original data",
        "Simulated, no floor",
        "Simulated, floor"
      )
    )
  )
  
  p <- ggplot2::ggplot(
    plot_df,
    ggplot2::aes(
      x = melanopic_edi_plot,
      y = emg_response_plot,
      colour = point_colour_group
    )
  ) +
    ggplot2::geom_point(alpha = 0.07) +
    ggplot2::geom_hline(
      yintercept = baseline_floor_emg,
      linetype = "dashed",
      linewidth = 0.8,
      colour = "black",
      alpha = 0.5
    ) +
    ggplot2::geom_text(
      data = baseline_label_df,
      ggplot2::aes(
        x = melanopic_edi_plot,
        y = emg_response_plot,
        label = label
      ),
      inherit.aes = FALSE,
      hjust = 1,
      vjust = -0.6,
      colour = "black",
      size = 3.5
    ) +
    ggplot2::geom_smooth(
      data = smooth_df,
      ggplot2::aes(
        group = smooth_group,
        colour = point_colour_group
      ),
      method = "lm",
      se = TRUE
    ) +
    ggplot2::scale_x_continuous(
      trans = LightLogR::symlog_trans(thr = 1),
      breaks = c(0, 1, 10, 100, 1000, 10000),
      minor_breaks = NULL
    ) +
    ggplot2::scale_y_log10(
      breaks = breaks,
      minor_breaks = minor_breaks
    ) +
    ggplot2::scale_colour_manual(
      values = c(
        "Original data" = "black",
        "Simulated, no floor" = "darkgreen",
        "Simulated, floor — not floored" = "darkgreen",
        "Simulated, floor — floored" = "darkred"
      ),
      breaks = c(
        "Original data",
        "Simulated, no floor",
        "Simulated, floor — not floored",
        "Simulated, floor — floored"
      ),
      labels = c(
        "Original data",
        "Simulated, no floor",
        "Simulated, floor: not floored",
        "Simulated, floor: floored"
      )
    ) +
    ggplot2::labs(
      x = "Melanopic EDI, lux",
      y = "OO-EMG response",
      colour = NULL,
      title = glue::glue(
        "Original model-fitting data vs simulated data; SD scaling factor = {sd_scaling}"
      )
    ) +
    ggplot2::theme_minimal()
  
  if (facet) {
    p <- p +
      ggplot2::facet_wrap(~ source, scales = "fixed") +
      ggplot2::guides(colour = "none")
  }
  
  print(p)
}

# Same plot, different colours
vis_sim_vs_original(sd_scaling = 1)

# Faceted version
vis_sim_vs_original(sd_scaling = 0.01, facet = TRUE)
```

# Collinearity of melanopic EDI and illuminance

We first look at collinearity of melanopic EDI and illuminance
If high collinearity, we cannot have both predictors in the same model

```{r}
#| fig-width: 8
#| fig-height: 12

design_matrix %>%
  dplyr::filter(MEDI > 0, illuminance > 0) %>%
  dplyr::summarise(
    cor_raw = cor(illuminance, MEDI),
    cor_log = cor(log10(illuminance), log10(MEDI)),
    sd_log_difference = sd(log10(illuminance) - log10(MEDI))
  )

# Simulate data
sim_dat <- simulate_from_design(design_df = design_matrix, model = m, sd_scaling = 1)

formula_model_col <- log10(simulated_emg) ~ log_illuminance + log_melanopic_edi + (1 | participant_id)

model_col <- lmer(formula_model_col, data = sim_dat)

summary(model_col)
performance::check_model(model_col)
cov2cor(vcov(model_col))
```

-> Correlation of 0.816 between the two predictors, which is too high.

# Model fit comparisons

```{r}
## Model fit comparisons with diagnostics
## glmmTMB only

fit_one_model <- function(formula, dat, model_name) {
  
  warnings <- character()
  error_msg <- NA_character_
  model <- NULL
  
  model <- tryCatch(
    withCallingHandlers(
      {
        glmmTMB::glmmTMB(
          formula = formula,
          data = dat,
          REML = FALSE
        )
      },
      warning = function(w) {
        warnings <<- c(warnings, conditionMessage(w))
        invokeRestart("muffleWarning")
      }
    ),
    error = function(e) {
      error_msg <<- conditionMessage(e)
      NULL
    }
  )
  
  if (is.null(model)) {
    return(
      list(
        model = NULL,
        diagnostics = tibble::tibble(
          model_name = model_name,
          converged = FALSE,
          convergence_code = NA_integer_,
          pdHess = NA,
          fit_message = NA_character_,
          warnings = paste(unique(warnings), collapse = " | "),
          error = error_msg,
          AIC = NA_real_
        )
      )
    )
  }
  
  convergence_code <- model$fit$convergence
  pdHess <- tryCatch(model$sdr$pdHess, error = function(e) NA)
  fit_message <- tryCatch(model$fit$message, error = function(e) NA_character_)
  
  converged <- isTRUE(convergence_code == 0) && isTRUE(pdHess)
  
  aic <- tryCatch(AIC(model), error = function(e) NA_real_)
  
  list(
    model = model,
    diagnostics = tibble::tibble(
      model_name = model_name,
      converged = converged,
      convergence_code = convergence_code,
      pdHess = pdHess,
      fit_message = fit_message,
      warnings = paste(unique(warnings), collapse = " | "),
      error = error_msg,
      AIC = aic
    )
  )
}

fit_models <- function(dat,
                       response = c("simulated_log10")) {
  
  response <- match.arg(response)
  
  dat <- dat %>%
    dplyr::mutate(
      participant_id = factor(participant_id),
      trial = as.numeric(trial),
      lights_on = as.logical(lights_on)
    )
  
  mel_formula <- stats::as.formula(
    paste(
      response,
      "~ log_melanopic_edi + log_baseline + trial +",
      "(log_melanopic_edi | participant_id)"
    )
  )
  
  phot_formula <- stats::as.formula(
    paste(
      response,
      "~ log_illuminance + log_baseline + trial +",
      "(log_illuminance | participant_id)"
    )
  )
  
  mel_fit <- fit_one_model(
    formula = mel_formula,
    dat = dat,
    model_name = "mel"
  )
  
  phot_fit <- fit_one_model(
    formula = phot_formula,
    dat = dat,
    model_name = "phot"
  )
  
  diagnostics <- dplyr::bind_rows(
    mel_fit$diagnostics,
    phot_fit$diagnostics
  )
  
  list(
    model_mel = mel_fit$model,
    model_phot = phot_fit$model,
    diagnostics = diagnostics
  )
}

compute_delta_aic <- function(model_list, delta_aic_threshold = 2) {
  
  diag <- model_list$diagnostics
  
  mel_ok <- diag %>%
    dplyr::filter(model_name == "mel") %>%
    dplyr::pull(converged)
  
  phot_ok <- diag %>%
    dplyr::filter(model_name == "phot") %>%
    dplyr::pull(converged)
  
  if (!isTRUE(mel_ok) || !isTRUE(phot_ok)) {
    return(
      tibble::tibble(
        AIC_mel = NA_real_,
        AIC_phot = NA_real_,
        delta_AIC = NA_real_,
        evidence = NA_character_
      )
    )
  }
  
  aic_mel <- diag %>%
    dplyr::filter(model_name == "mel") %>%
    dplyr::pull(AIC)
  
  aic_phot <- diag %>%
    dplyr::filter(model_name == "phot") %>%
    dplyr::pull(AIC)
  
  delta_aic <- aic_mel - aic_phot
  
  evidence <- dplyr::case_when(
    delta_aic < -delta_aic_threshold ~ "favours melanopic EDI model",
    delta_aic >  delta_aic_threshold ~ "favours photopic lux model",
    TRUE                             ~ "no clear preference"
  )
  
  tibble::tibble(
    AIC_mel = aic_mel,
    AIC_phot = aic_phot,
    delta_AIC = delta_aic,
    evidence = evidence
  )
}
```

# BFDA loop

```{r warning=F}
future::plan(
  future::multisession,
  workers = parallel::detectCores() - 1
)

run_aic_sim <- function(
    n_sims = 1000,
    n_subjects = 20,
    noise_scaling = 1,
    delta_aic_threshold = 2
) {
  
  response <- "simulated_log10"
  
  available_subjects <- design_matrix %>%
    dplyr::distinct(participant_id) %>%
    dplyr::arrange(participant_id) %>%
    dplyr::pull(participant_id)
  
  if (n_subjects > length(available_subjects)) {
    stop(
      "Requested ", n_subjects, " participants, but design_matrix only contains ",
      length(available_subjects), ".",
      call. = FALSE
    )
  }
  
  selected_subjects <- available_subjects[seq_len(n_subjects)]
  
  design_subset <- design_matrix %>%
    dplyr::filter(participant_id %in% selected_subjects) %>%
    dplyr::mutate(
      participant_id = factor(participant_id)
    )
  
  sim_results <- furrr::future_map(
    seq_len(n_sims),
    ~ {
      sim_data <- simulate_from_design(
        design_df = design_subset,
        model = m,
        sd_scaling = noise_scaling
      )
      
      fits <- fit_models(
        dat = sim_data,
        response = response
      )
      
      aic_result <- compute_delta_aic(
        fits,
        delta_aic_threshold = delta_aic_threshold
      )
      
      data_diagnostics <- tibble::tibble(
        n_obs = nrow(sim_data),
        n_participants = dplyr::n_distinct(sim_data$participant_id),
        n_lab_visits = dplyr::n_distinct(sim_data$lab_visit),
        sd_response = stats::sd(sim_data[[response]], na.rm = TRUE),
        sd_log_melanopic_edi = stats::sd(sim_data$log_melanopic_edi, na.rm = TRUE),
        sd_log_illuminance = stats::sd(sim_data$log_illuminance, na.rm = TRUE),
        prop_floored = mean(sim_data$floored, na.rm = TRUE)
      )
      
      diagnostics <- fits$diagnostics %>%
        dplyr::mutate(
          sim_id = .x,
          n_subjects = n_subjects,
          noise_scaling = noise_scaling,
          response = response
        ) %>%
        dplyr::bind_cols(
          data_diagnostics[rep(1, nrow(.)), ]
        )
      
      list(
        aic_result = aic_result %>%
          dplyr::mutate(
            sim_id = .x,
            n_subjects = n_subjects,
            noise_scaling = noise_scaling,
            response = response
          ),
        diagnostics = diagnostics
      )
    },
    .options = furrr::furrr_options(seed = TRUE)
  )
  
  aic_results <- purrr::map_dfr(sim_results, "aic_result")
  diagnostics <- purrr::map_dfr(sim_results, "diagnostics")
  
  delta_aic_list <- aic_results$delta_AIC
  
  n_valid <- sum(!is.na(delta_aic_list))
  n_failed <- sum(is.na(delta_aic_list))
  
  if (n_valid == 0) {
    summary <- tibble::tibble(
      p_favour_mel = NA_real_,
      p_unclear = NA_real_,
      p_favour_phot = NA_real_,
      n_sims = n_sims,
      n_valid = n_valid,
      n_failed = n_failed,
      prop_failed = n_failed / n_sims,
      n_subjects = n_subjects,
      noise_scaling = noise_scaling,
      response = response
    )
  } else {
    summary <- tibble::tibble(
      p_favour_mel = mean(delta_aic_list < -delta_aic_threshold, na.rm = TRUE),
      p_unclear = mean(abs(delta_aic_list) <= delta_aic_threshold, na.rm = TRUE),
      p_favour_phot = mean(delta_aic_list > delta_aic_threshold, na.rm = TRUE),
      n_sims = n_sims,
      n_valid = n_valid,
      n_failed = n_failed,
      prop_failed = n_failed / n_sims,
      n_subjects = n_subjects,
      noise_scaling = noise_scaling,
      response = response
    )
  }
  
  list(
    summary = summary,
    aic_results = aic_results,
    diagnostics = diagnostics
  )
}

# grid <- tidyr::crossing(
#   n_subjects = as.integer(seq(2, 50)), # 2 to 50 participants
#   noise_scaling = as.integer(seq(1, 10)) # 10 values, range is from original residual variance to residual variance increased 10 fold
# )

grid <- tidyr::crossing(
  n_subjects = c(2, 5, 10, 15, 20, 30, 40, 50),
  noise_scaling = c(0.1, 0.5, 1, 2, 5)
)

aic_nested <- grid %>%
  dplyr::mutate(
    results = purrr::pmap(
      list(n_subjects, noise_scaling),
      ~ {
        cat(
          "Running: subjects =", ..1,
          " noise =", ..2,
          "\n"
        )
        
        run_aic_sim(
          n_sims = 20,
          n_subjects = ..1,
          noise_scaling = ..2
        )
      }
    )
  )

aic_df <- aic_nested %>%
  dplyr::mutate(
    summary = purrr::map(
      results,
      ~ .x$summary %>%
        dplyr::select(
          -dplyr::any_of(c("n_subjects", "noise_scaling"))
        )
    )
  ) %>%
  dplyr::select(-results) %>%
  tidyr::unnest(summary)

aic_diagnostics <- aic_nested %>%
  dplyr::mutate(
    diagnostics = purrr::map(
      results,
      ~ .x$diagnostics %>%
        dplyr::select(
          -dplyr::any_of(c("n_subjects", "noise_scaling"))
        )
    )
  ) %>%
  dplyr::select(-results) %>%
  tidyr::unnest(diagnostics)

# aic_diagnostics %>%
#   dplyr::group_by(noise_scaling, model_name) %>%
#   dplyr::summarise(
#     n_fits = dplyr::n(),
#     n_failed = sum(!converged),
#     prop_failed = mean(!converged),
#     convergence_codes = paste(unique(convergence_code), collapse = ", "),
#     pdHess_values = paste(unique(pdHess), collapse = ", "),
#     warnings = paste(unique(warnings[warnings != ""]), collapse = " | "),
#     errors = paste(unique(error[!is.na(error)]), collapse = " | "),
#     .groups = "drop"
#   ) %>%
#   dplyr::arrange(noise_scaling, model_name)

# aic_diagnostics %>%
#   dplyr::group_by(noise_scaling, n_subjects, response, model_name) %>%
#   dplyr::summarise(
#     n_fits = dplyr::n(),
#     n_failed = sum(!converged),
#     prop_failed = mean(!converged),
#     pdHess_values = paste(unique(pdHess), collapse = ", "),
#     warnings = paste(unique(warnings[warnings != ""]), collapse = " | "),
#     fit_messages = paste(unique(fit_message[!is.na(fit_message)]), collapse = " | "),
#     errors = paste(unique(error[!is.na(error)]), collapse = " | "),
#     mean_prop_floored = mean(prop_floored, na.rm = TRUE),
#     .groups = "drop"
#   )

aic_diagnostics %>%
  dplyr::group_by(n_subjects, noise_scaling, model_name) %>%
  dplyr::summarise(
    n_fits = dplyr::n(),
    n_failed = sum(!converged),
    prop_failed = mean(!converged),
    warnings = paste(unique(warnings[warnings != ""]), collapse = " | "),
    fit_messages = paste(unique(fit_message[!is.na(fit_message)]), collapse = " | "),
    errors = paste(unique(error[!is.na(error)]), collapse = " | "),
    .groups = "drop"
  )

aic_per_sim <- aic_nested %>%
  dplyr::mutate(
    aic_results = purrr::map(results, "aic_results")
  ) %>%
  dplyr::select(aic_results) %>%
  tidyr::unnest(aic_results)
```

```{r}
readr::write_csv(
  aic_df,
  here::here("outputs", "aic_sim_results.csv")
)

readr::write_csv(
  aic_diagnostics,
  here::here("outputs", "aic_sim_diagnostics.csv")
)

readr::write_csv(
  aic_per_sim,
  here::here("outputs", "aic_sim_per_simulation_results.csv")
)
```


# Plotting results

```{r}
# aic_df <- read_csv(here("outputs", "aic_sim_results.csv"))

power_threshold <- 0.8

# Use probability of selecting the melanopic model as the main "power" outcome
plot_df <- aic_df %>%
  dplyr::mutate(
    power = p_favour_mel
  )

# Colour palettes 
n_levels_noise <- length(levels(factor(plot_df$noise_scaling))) 
n_levels_subjects <- length(levels(factor(plot_df$n_subjects))) 
plot1_pal <- colorRampPalette(c("#E69F00", "#2C3E50")) 
plot2_pal <- colorRampPalette(c("#2C3E50", "#E69F00")) 
plot3_pal <- colorRampPalette(c("#F2C14E", "#8E1B1B"))

# 1. Power vs number of participants
## This shows how power increases with participants, with separate curves for noise levels.
plot1 <- ggplot(
  plot_df,
  aes(x = n_subjects, y = power, group = factor(noise_scaling), colour = factor(noise_scaling))
) + 
  geom_line() +
  geom_point() + 
  
  # ---- Annotation text ----
  annotate("text",
           x = 35,
           y = 0.115,
           label = "Selected sample size",
           colour = "black",
           vjust = 0,
           size = 3, 
           fontface = "bold") +
  
  # ---- Left-facing arrow below text ----
  annotate("segment",
           x = 27, xend = 22,       # ← makes arrow point left
           y = 0.125, yend = 0.125,   # position slightly below text
           colour = "black",
           arrow = arrow(type = "closed", length = unit(0.15, "cm"))) +
  
  annotate("text",
           x = 40,
           y = power_threshold + 0.02,
           label = "Power threshold",
           colour = "red",
           vjust = 0, 
           size = 3, 
           fontface = "bold") +
  
  geom_hline(yintercept = power_threshold, color = "red", linetype = "dashed") + 
  geom_vline(xintercept = 20, colour = "black", linetype = "dashed", linewidth = 1) +
  scale_colour_manual(values = plot1_pal(n_levels_noise)) +
  xlim(2, 50) +
  ylim(0, 1) + 
  labs(x = "Sample size", 
       y = "Power: P(ΔAIC < -2)",
       colour = "Noise scaling factor", 
       title = "**A** Dependence of AIC-based power on sample size") + 
  theme_minimal() + 
  theme(
    plot.title = ggtext::element_markdown(),
    legend.key.height = unit(3, "mm"),     # vertical size of each key
    legend.key.width  = unit(4, "mm"),     # optional, horizontal
    legend.spacing.y  = unit(1, "mm")      # space between legend rows
  ) +   
  plot_formatting

plot20 <- plot_df %>% dplyr::filter(n_subjects == 20)

# 2. Power vs noise scaling
## Because noise scaling is on a log10 scale, we apply scale_x_log10().
plot2 <- ggplot(plot_df, aes(x = noise_scaling, y = power, group = n_subjects, colour = n_subjects)) + 
  geom_line() + 
  geom_point() + 

  # --- threshold line (existing) ---
  geom_hline(yintercept = power_threshold, color = "red", linetype = "dashed") +

  # --- label above horizontal threshold ---
  annotate("text",
           x = 8,
           y = power_threshold + 0.02,
           label = "Power threshold",
           colour = "red",
           vjust = 0,
           size = 3, 
           fontface = "bold") +

  # --- selected sample size line ---
  geom_line(data = plot20,
            aes(x = noise_scaling, y = power),
            colour = "black",
            linetype = "dashed",
            linewidth = 1,
            inherit.aes = FALSE) +
  
  annotation_custom(
    grob = grid::textGrob(
      label = "– –  Selected sample size",
      x = 0.95, y = 0.98,              # right-aligned, above plot
      hjust = 1, vjust = 0,
      gp = grid::gpar(col = "black", fontsize = 8, fontface = "bold")
    )
  ) +

  scale_colour_gradientn(colors = plot2_pal(n_levels_subjects)) +
  scale_x_continuous() +
  coord_cartesian(clip = "off") +
  ylim(0, 1) +
  labs(x = "Noise scaling factor",
       y = "Power: P(ΔAIC < -2)",
       colour = "Sample size",
       title = "**B** Dependence of AIC-based power on noise scaling") +
  theme_minimal() +
  theme(
    plot.title = ggtext::element_markdown()
    ) +     
  plot_formatting

# 3. Power heatmap (matrix)
## Power-analysis view: participants × noise scaling. This gives a nice heatmap of the whole design grid.

plot_df <- plot_df %>%
  dplyr::mutate(
    y_index = match(n_subjects, sort(unique(n_subjects))),
    x_index = match(noise_scaling, sort(unique(noise_scaling)))
  )

plot3 <- ggplot(plot_df, aes(x = x_index, y = y_index, fill = power)) + 
  geom_tile() +
  geom_hline(
    yintercept = match(20, sort(unique(plot_df$n_subjects))),
    colour = "black", linetype = "dashed", linewidth = 1
  ) +
  geom_text(
  aes(
    label = sprintf("%.2f", power),
    fontface = ifelse(power >= power_threshold, "bold", "plain")
  ),
  colour = "white",
  size = 3
) +
  scale_fill_gradientn(colours = plot3_pal(100), guide = "none") +
  scale_x_continuous(
    breaks = unique(plot_df$x_index),
    labels = sort(unique(plot_df$noise_scaling))
  ) +
  scale_y_continuous(
    breaks = unique(plot_df$y_index),
    labels = sort(unique(plot_df$n_subjects))
  ) +
  labs(
    x = "Noise scaling factor", 
    y = "Sample size",
    title = "AIC-based power heatmap"
  ) +
  annotation_custom(
    grob = grid::textGrob(
      label = "– –  Selected sample size",
      x = 0.95, y = 0.98,              # right-aligned, above plot
      hjust = 1, vjust = 0,
      gp = grid::gpar(col = "black", fontsize = 8, fontface = "bold")
    )
  ) +
  coord_cartesian(clip = "off") +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    axis.ticks = element_blank()
  ) +
  plot_formatting

joined <- plot1 / plot2
# + plot_annotation(title = "Power analysis (BFDA)")

joined
plot1
plot2
plot3

# Saving plots
save_rds_safe(plot1, here("outputs", "figures", "rds", glue("aic_power_plot_a_sample_size.rds")))
save_rds_safe(plot2, here("outputs", "figures", "rds", glue("aic_power_plot_b_noise_scaling.rds")))

formatted_save(here("outputs", "figures", "aic_power_plot_ab_combined.jpg"), joined, width = 7, height = 6)
save_rds_safe(joined, here("outputs", "figures", "rds", glue("aic_power_plot_ab_combined.rds")))

save_rds_safe(plot3, here("outputs", "figures", "rds", glue("aic_power_plot_c_heatmap.rds")))
```

# Contour plot

```{r}
plot3_lines <- ggplot(plot_df, aes(noise_scaling, n_subjects, z = power)) +
  geom_contour(bins = 10, linewidth = 0.4, colour = "grey30") +
  geom_contour(breaks = power_threshold, colour = "red", linewidth = 1) +
  geom_hline(yintercept = 20, colour = "black", linetype = "dashed", linewidth = 1) +
  scale_x_continuous(breaks = seq(1, 10)) +
  labs(x = "Noise scaling factor", y = "Sample size", title = "AIC-based power contours") +
  coord_cartesian(clip = "off") +
  theme_minimal() +
  theme(panel.grid = element_blank(), axis.ticks = element_blank()) +
  plot_formatting

plot3_lines

# -------

contour_breaks <- c(seq(0, 1, by = 0.1), 1.000001)

nice_interval_labels <- function(x) {
  # Round anything like 0.700000 -> 0.7 (to 1 decimal place)
  x <- gsub("(\\d+\\.\\d{1})\\d+", "\\1", x)

  # Rename the top bin (after rounding it becomes "(1, 1.0]")
  x <- gsub("^\\(1(\\.0)?,\\s*1(\\.0)?\\]$", "1.0", x)

  x
}


plot3_contour <- ggplot(
  plot_df,
  aes(
    x = noise_scaling,
    y = n_subjects,
    z = power
  )
) +
  geom_contour_filled(
    breaks = contour_breaks,
    alpha = 0.95
  ) +
  
  geom_contour(
    breaks = power_threshold,
    colour = "red",
    linewidth = 1
  ) +
  
  # geom_raster(aes(fill = power), interpolate = TRUE) +

  # --- selected sample size line (n = 20) ---
  geom_hline(
    yintercept = 20,
    colour = "black",
    linetype = "dashed",
    linewidth = 1
  ) +

  scale_fill_manual(
    values = plot3_pal(11),
    name = "Power: P(ΔAIC < -2)",
    labels = nice_interval_labels
  ) +
  
  scale_x_continuous(breaks = seq(1, 10)) +

  labs(
    x = "Noise scaling factor",
    y = "Sample size",
    title = "AIC-based power surface"
  ) +

  annotation_custom(
    grob = grid::textGrob(
      label = "– –  Selected sample size",
      x = 0.95, y = 0.98,
      hjust = 1, vjust = 0,
      gp = grid::gpar(col = "black", fontsize = 8, fontface = "bold")
    )
  ) +

  coord_cartesian(clip = "off") +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    axis.ticks = element_blank()
  ) +
  plot_formatting

plot3_contour

formatted_save(here("outputs", "figures", "aic_power_plot_c_contour.jpg"), plot3_contour)
save_rds_safe(plot3_contour, here("outputs", "figures", "rds", glue("aic_power_plot_c_contour.rds")))
```

Extra plot…

```{r}
iso_breaks <- seq(0, 1, by = 0.1)

plot3_iso_grad <- ggplot(
  plot_df,
  aes(x = noise_scaling, y = n_subjects, z = power)
) +
  geom_contour(
    aes(colour = after_stat(level)),
    breaks = iso_breaks,
    linewidth = 0.8
  ) +
  
  geom_hline(
    yintercept = 20,
    colour = "black",
    linetype = "dashed",
    linewidth = 1
  ) +
  
  scale_colour_gradientn(
    colours = plot3_pal(100),
    name = "Power: P(ΔAIC < -2)",
    limits = c(0, 1),
    breaks = iso_breaks,
    labels = function(x) sprintf("%.1f", x)
  ) +
  
  scale_x_continuous(breaks = seq(1, 10)) +
  
  labs(
    x = "Noise scaling factor",
    y = "Sample size",
    title = "AIC-based power contour plot"
  ) +
  
  annotation_custom(
    grob = grid::textGrob(
      label = "– –  Selected sample size",
      x = 0.95,
      y = 0.98,
      hjust = 1,
      vjust = 0,
      gp = grid::gpar(
        col = "black",
        fontsize = 8,
        fontface = "bold"
      )
    )
  ) +
  
  coord_cartesian(clip = "off") +
  theme_minimal() +
  plot_formatting

plot3_iso_grad

save_rds_safe(
  plot3_iso_grad,
  here::here("outputs", "figures", "rds", glue::glue("aic_power_plot_c_contour_bis.rds"))
)
```

```{r}
sessionInfo()
```

```{r render-time-2, echo=FALSE}
render_time <- Sys.time() - .render_start
cat("Render time:", format(render_time))
```
