---
title: "Generate design matrix"
author: "Lucien Bickerstaff"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
---

```{r render-time-1, include=FALSE}
# Record how long it took to render the file
.render_start <- Sys.time()
```

```{r knitr-options, include=FALSE}
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE
)
```

# Setup

```{r setup, warning=FALSE, message=FALSE}
# Seed for reproducibility
set.seed(2025)

# Libraries
library(here)
library(tidyverse)
library(openxlsx)
library(photobiology)
library(grid)
library(patchwork)
library(glue)

source(here("R", "helpers.R"))
source(here("R", "plot_settings.R"))
```

# Function to preprocess the spectral data

```{r export-jeti-spectra}
export_jeti_spectra_csv <- function(input_xlsx,
                                    output_csv,
                                    sheet = 1,
                                    name_col = "Name",
                                    wl_min = 380,
                                    wl_max = 780,
                                    measurement_prefix = "measurement_") {
  # Read raw Excel file without headers
  raw <- openxlsx::read.xlsx(input_xlsx, sheet = sheet, colNames = FALSE)
  
  # Transpose so first row becomes header row
  dat <- as.data.frame(t(raw), stringsAsFactors = FALSE)
  colnames(dat) <- dat[1, ]
  dat <- dat[-1, , drop = FALSE]
  
  # Define wavelength columns expected in the file
  wl_cols <- as.character(wl_min:wl_max)
  
  # Check required columns exist
  missing_cols <- setdiff(c(name_col, wl_cols), names(dat))
  if (length(missing_cols) > 0) {
    stop(
      "Missing expected columns: ",
      paste(missing_cols, collapse = ", ")
    )
  }
  
  # Keep only Name + spectral columns
  spectral_data <- dat %>%
    select(all_of(c(name_col, wl_cols))) %>%
    pivot_longer(
      cols = all_of(wl_cols),
      names_to = "wavelength",
      values_to = "value"
    ) %>%
    mutate(
      wavelength = as.integer(wavelength),
      value = as.numeric(gsub(",", ".", value, fixed = TRUE)),
      measurement = str_extract(.data[[name_col]], "(?<=#)\\d+")
    ) %>% 
    dplyr::filter(!is.na(measurement))
  
  # Fall back to original row name if no #number pattern is found
  if (all(is.na(spectral_data$measurement))) {
    spectral_data <- spectral_data %>%
      mutate(measurement = .data[[name_col]])
  }
  
  # Order measurements naturally if they are numeric IDs
  measurement_levels <- spectral_data %>%
    distinct(measurement) %>%
    mutate(measurement_num = suppressWarnings(as.numeric(as.character(measurement)))) %>%
    arrange(measurement_num, measurement) %>%
    filter(!is.na(measurement)) %>%
    pull(measurement)
  
  spectral_data <- spectral_data %>%
    mutate(measurement = factor(measurement, levels = measurement_levels))
  
  # Convert to wide format: wavelength in first column, measurements in remaining columns
  out <- spectral_data %>%
    transmute(
      wavelength,
      measurement = paste0(measurement_prefix, measurement),
      value
    ) %>%
    pivot_wider(
      names_from = measurement,
      values_from = value,
      values_fn = \(x) x[1]
    ) %>%
    arrange(wavelength)
  
  meas_cols <- setdiff(names(out), "wavelength")
  out <- out %>%
    select(
      wavelength,
      all_of(meas_cols[order(as.numeric(sub("^measurement_", "", meas_cols)))])
    )
  
  # Rename columns
  names(out) <- c(
    "Wavelength",
    paste0("Measurement ", seq_len(ncol(out) - 1))
  )
  
  # Format values to avoid scientific notation (keep numeric-looking strings)
  out_export <- out %>%
    mutate(across(
      -Wavelength,
      ~ format(.x, scientific = FALSE, digits = 10)
    ))
  
  # Write CSV with headers
  readr::write_csv(out_export, output_csv, na = "")

  # Return data invisibly for immediate use in R
  return(invisible(out))
}

export_jeti_spectra_csv(
  input_xlsx = here::here("data", "light_source_calibration", "1_raw", "1_calibration_reference_raw.xlsx"),
  output_csv = here::here("data", "light_source_calibration", "2_preprocessed", "1_calibration_reference_preprocessed.csv")
)

export_jeti_spectra_csv(
  input_xlsx = here::here("data", "light_source_calibration", "1_raw", "2a_target_settings_raw.xlsx"),
  output_csv = here::here("data", "light_source_calibration", "2_preprocessed", "2a_target_settings_preprocessed.csv")
)

export_jeti_spectra_csv(
  input_xlsx = here::here("data", "light_source_calibration", "1_raw", "2b_target_settings_corrected_raw.xlsx"),
  output_csv = here::here("data", "light_source_calibration", "2_preprocessed", "2b_target_settings_corrected_preprocessed.csv")
)

export_jeti_spectra_csv(
  input_xlsx = here::here("data", "light_source_calibration", "1_raw", "3_pilot_settings_raw.xlsx"),
  output_csv = here::here("data", "light_source_calibration", "2_preprocessed", "3_pilot_settings_preprocessed.csv")
)
```

The preprocessed data are then passed to Luox to calculate alpha-opic quantities. This moves the data from step 2 (preprocessed spectra) to step 3 (Luox outputs).

We then extract the relevant outcomes from step 3:

```{r prepare-luox-outcomes}
prepare_outcomes <- function(
  file_stub,
  luox_suffix = "_luox.csv",
  raw_suffix = "_raw.xlsx",
  irradiance_label = "Ee [W/(sqm)]"
) {
  
  luox_file <- here::here(
    "data", "light_source_calibration", "3_luox_outputs",
    paste0(file_stub, luox_suffix)
  )
  raw_file <- here::here(
    "data", "light_source_calibration", "1_raw",
    paste0(file_stub, raw_suffix)
  )
  
  if (!file.exists(luox_file)) {
    stop("Luox file not found: ", luox_file)
  }
  if (!file.exists(raw_file)) {
    stop("Raw file not found: ", raw_file)
  }
  
  # -----------------------------
  # 1) Read luox file and reshape
  # -----------------------------
  lux_data_wide <- readr::read_csv(
    luox_file,
    show_col_types = FALSE,
    col_types = readr::cols(.default = readr::col_character())
  )
  
  first_col <- names(lux_data_wide)[1]
  measurement_cols <- grep("^Measurement\\s*\\d+$", names(lux_data_wide), value = TRUE)
  
  if (length(measurement_cols) == 0) {
    stop("No columns like 'Measurement 1', 'Measurement 2', ... found in luox file.")
  }
  
  lux_data <- lux_data_wide %>%
    tidyr::pivot_longer(
      cols = dplyr::all_of(measurement_cols),
      names_to = "measurement_label",
      values_to = "value"
    ) %>%
    dplyr::mutate(
      value = as.numeric(gsub(",", ".", value)),
      measurement = stringr::str_extract(measurement_label, "\\d+"),
      measurement = factor(measurement)
    ) %>%
    tidyr::pivot_wider(
      names_from = dplyr::all_of(first_col),
      values_from = value
    ) %>% 
    select(-measurement_label)
  
  # -----------------------------
  # 2) Read and transpose raw file
  #    mirroring your reference code
  # -----------------------------
  data <- openxlsx::read.xlsx(raw_file, colNames = FALSE)
  data <- as.data.frame(t(data), stringsAsFactors = FALSE)
  colnames(data) <- data[1, ]
  data <- data[-1, , drop = FALSE]
  
  # -----------------------------
  # 3) Extract spectral data
  # -----------------------------
  spectral_data <- data %>%
    dplyr::select(Name, `380`:`780`)
  
  spectral_data <- spectral_data %>%
    tidyr::pivot_longer(
      cols = -1,
      names_to = "wavelength",
      values_to = "Ee [W/(sqm*nm)]"
    ) %>%
    dplyr::mutate(
      dplyr::across(
        -1,
        ~ as.numeric(gsub(",", ".", .))
      )
    ) %>%
    dplyr::mutate(
      measurement = stringr::str_extract(Name, "(?<=#)\\d+"),
      measurement = factor(measurement)
    ) %>%
    dplyr::mutate(
      measurement = forcats::fct_reorder(
        measurement,
        as.numeric(as.character(measurement))
      )
    ) %>%
    dplyr::select(-Name)
  
  # -----------------------------
  # 4) Calculate irradiance
  # -----------------------------
  irradiance_data <- spectral_data %>%
    dplyr::group_by(measurement) %>%
    dplyr::summarise(
      !!irradiance_label := sum(`Ee [W/(sqm*nm)]`, na.rm = TRUE),
      .groups = "drop"
    )
  
  # -----------------------------
  # 5) Merge irradiance into luox/outcomes
  # -----------------------------
  merged_data <- dplyr::left_join(
    irradiance_data,
    lux_data,
    by = c("measurement")
  )
  
  return(merged_data)
}
```

# Prepare calibration data

```{r prepare-calibration-data}
calibration <- prepare_outcomes("1_calibration_reference")

calibration <- calibration %>% 
  mutate(measurement_num = as.numeric(measurement))

# -----------------------------
# Add primary mapping
# -----------------------------
calibration <- calibration %>%
  dplyr::mutate(
    primary = dplyr::case_when(
      measurement_num >= 1   & measurement_num <= 15  ~ "1",
      measurement_num >= 16  & measurement_num <= 30  ~ "2",
      measurement_num >= 31  & measurement_num <= 45  ~ "3",
      measurement_num >= 46  & measurement_num <= 60  ~ "4",
      measurement_num >= 61  & measurement_num <= 75  ~ "5",
      measurement_num >= 76  & measurement_num <= 90  ~ "6",
      measurement_num >= 91  & measurement_num <= 105 ~ "7",
      measurement_num >= 106 & measurement_num <= 120 ~ "8",
      measurement_num >= 121 & measurement_num <= 135 ~ "9",
      measurement_num >= 136 & measurement_num <= 150 ~ "10",
      measurement_num >= 151 & measurement_num <= 165 ~ "all",
      TRUE ~ NA_character_
    ),
    primary = factor(primary, levels = c(as.character(1:10), "all"))
  ) %>%
  dplyr::filter(!is.na(primary))

# -----------------------------
# Add measurement-within-primary index
# -----------------------------
calibration <- calibration %>%
  dplyr::group_by(primary) %>%
  dplyr::mutate(meas_within_primary = dplyr::row_number()) %>%
  dplyr::ungroup()

# -----------------------------
# Map settings
# -----------------------------
settings_vec <- c(
  25, 50, 100, 200, 400, 800, 1200, 1600, 2000,
  2400, 2800, 3200, 3600, 4000, 4095
)

calibration <- calibration %>%
  mutate(setting = settings_vec[meas_within_primary]) %>% 
  rename(
    irradiance = `Ee [W/(sqm)]`,
    illuminance = `Illuminance (lx)`,
    mel_edi = `Melanopic EDI (lx)`,
    rho_edi = `Rhodopic EDI (lx)`
  ) %>% 
  select(
    measurement_num,
    primary,
    meas_within_primary,
    setting,
    irradiance,
    illuminance,
    mel_edi,
    rho_edi
  )

# -----------------------------
# Write output
# -----------------------------
readr::write_csv(
  calibration, 
  here("data", "light_source_calibration", "4_outcomes", "1_calibration_reference_outcomes.csv"), 
  na = ""
)
```

## Control plots

```{r plot-calibration-control}
ggplot(data=calibration, aes(x=setting, y=irradiance)) +
  geom_point() +
  labs(title = "Irradiance") +
  facet_wrap(~ primary) +
  plot_formatting +
  theme_minimal()
ggplot(data=calibration, aes(x=setting, y=illuminance)) +
  geom_point() +
  labs(title = "Photopic illuminance") +
  facet_wrap(~ primary) +
  plot_formatting +
  theme_minimal()
ggplot(data=calibration, aes(x=setting, y=mel_edi)) +
  geom_point() +
  labs(title = "Melanopic EDI") +
  facet_wrap(~ primary) +
  plot_formatting +
  theme_minimal()
ggplot(data=calibration, aes(x=setting, y=rho_edi)) +
  geom_point() +
  labs(title = "Rhodopic EDI") +
  facet_wrap(~ primary) +
  plot_formatting +
  theme_minimal()
```

# Find max possible irradiance for each primary of interest

```{r find-primary-max-irradiance, echo=TRUE, fold=FALSE}
calibration_select <- calibration %>%
  dplyr::filter(primary %in% as.character(2:7))

max_irradiance_data <- calibration_select %>%
  group_by(primary) %>%
  summarise(
    max_irr = max(irradiance, na.rm = TRUE)
  )

# The max possible irradiance value is the min irradiance value achievable
# by all primaries overall
min_irradiance <- min(max_irradiance_data$max_irr)
min_irradiance
```

This is the maxmimum irradiance we can achieve across all primaries.

# Define 4 log-spaced target irradiances 

Four target irradiance values for each of the four lab visits

```{r define-target-irradiances, echo=TRUE, fold=FALSE}
decrease_factor <- 150 # how much smaller the min irradiance value is compared to the max value

max_val <- min_irradiance
min_val <- max_val / decrease_factor
n_values <- 4

irradiance_levels <- exp(seq(log(max_val), log(min_val), length.out = n_values))
irradiance_levels
```

# Interpolation of calibration data

We interpolate the data so we have a [setting] -> [intensity metric] map

```{r interpolate-primary-settings, echo=TRUE, fold=FALSE}
# Define target settings (full interpolation range)
all_settings <- 0:4095  # start from 0 now

interp_data <- calibration %>%
  group_by(primary) %>%
  group_modify(~ {
    df <- .x[order(.x$setting), ]

    # Base interpolation range (0 to 4095)
    xout <- all_settings

    # --- Irradiance ---
    Ee_fit <- spline(
      x = df$setting,
      y = df$irradiance,
      xout = xout
    )$y

    # --- Photopic illuminance ---
    Ev_fit <- spline(
      x = df$setting,
      y = df$illuminance,
      xout = xout
    )$y

    # --- Melanopic EDI ---
    MEDI_fit <- spline(
      x = df$setting,
      y = df$mel_edi,
      xout = xout
    )$y

    # --- Rhodopic EDI ---
    REDI_fit <- spline(
      x = df$setting,
      y = df$rho_edi,
      xout = xout
    )$y

    # --- Linear extension to 0 ---
    first_x <- df$setting[1]
    first_Ee <- df$irradiance[1]
    first_Ev <- df$illuminance[1]
    first_MEDI <- df$mel_edi[1]
    first_REDI <- df$rho_edi[1]

    Ee_fit[xout < first_x] <- (xout[xout < first_x] / first_x) * first_Ee
    Ev_fit[xout < first_x] <- (xout[xout < first_x] / first_x) * first_Ev
    MEDI_fit[xout < first_x] <- (xout[xout < first_x] / first_x) * first_MEDI
    REDI_fit[xout < first_x] <- (xout[xout < first_x] / first_x) * first_REDI

    tibble(
      setting = xout,
      Ee = Ee_fit,
      Ev = Ev_fit,
      MEDI = MEDI_fit,
      REDI = REDI_fit
    )
  }) %>%
  ungroup()
```

## Control plots

```{r plot-interpolation-control}
# Irradiance
ggplot(interp_data, aes(x = setting, y = Ee, color = primary)) +
  geom_line() +
  geom_point(data = calibration, aes(y = irradiance), color = "black", size = 0.8) +
  theme_minimal() +
  facet_wrap(~ primary, scales = "free") +
  labs(y = "Irradiance (W/m²)", title = "Interpolated Irradiance per Primary") +
  plot_formatting

# Illuminance
ggplot(interp_data, aes(x = setting, y = Ev, color = primary)) +
  geom_line() +
  geom_point(data = calibration, aes(y = illuminance), color = "black", size = 0.8) +
  theme_minimal() +
  facet_wrap(~ primary, scales = "free") +
  labs(y = "Ev [lx] (CIE1931 2°)", title = "Interpolated Photopic Illuminance per Primary") +
  plot_formatting

# Melanopic EDI
ggplot(interp_data, aes(x = setting, y = MEDI, color = primary)) +
  geom_line() +
  geom_point(data = calibration, aes(y = mel_edi), color = "black", size = 0.8) +
  theme_minimal() +
  facet_wrap(~ primary, scales = "free") +
  labs(y = "X v,mel,D65 (MEDI) [lx]", title = "Interpolated Melanopic EDI per Primary") +
  plot_formatting

# Scotopic lux
ggplot(interp_data, aes(x = setting, y = REDI, color = primary)) +
  geom_line() +
  geom_point(data = calibration, aes(y = rho_edi), color = "black", size = 0.8) +
  theme_minimal() +
  facet_wrap(~ primary, scales = "free") +
  labs(y = "Rhodopic EDI [lx]", title = "Interpolated Rhodopic EDI per Primary") +
  plot_formatting
```

# Map target irradiance levels to needed settings for each primary

Now, we just map our target irradiance levels to the desired settings…

```{r map-targets-to-settings, warning=FALSE, echo=TRUE, fold=FALSE}
interp_data_all <- interp_data # keeping old df…

interp_data <- interp_data %>% 
  dplyr::filter(primary != "all") %>%
  dplyr::filter(primary %in% as.character(2:7))

mapped_settings <- calibration

# For each primary and each target irradiance, find the closest setting
mapped_settings <- expand_grid(
  primary = unique(interp_data$primary),
  target_irr = irradiance_levels
) %>%
  left_join(interp_data, by = "primary") %>%
  group_by(primary, target_irr) %>%
  summarise(
    # Find index of nearest irradiance to the target
    idx = which.min(abs(Ee - target_irr)),

    # Extract corresponding setting and all interpolated modalities
    setting = setting[idx],
    irradiance_theoretical = Ee[idx],
    illuminance_theoretical = Ev[idx],
    MEDI_theoretical = MEDI[idx],
    REDI_theoretical = REDI[idx],
    .groups = "drop"
  ) %>%
  group_by(primary) %>%
  mutate(
    lab_visit = dense_rank(target_irr),  # 1 = lowest, 4 = highest
    lab_visit = factor(lab_visit, levels = 1:4, labels = c("1", "2", "3", "4"))
  ) %>%
  ungroup() %>% 
  select(-idx)
```

# Add peak wavelength information

Extracted from Spectra Tune Lab manual

```{r add-peak-wavelengths}
peak_nm_df <- data.frame(primary = factor(c(1:10)),
                         peak_nm = c(430, 445, 465, 475, 505, 525, 545, 595, 640, 660),
                         color = c("UV", "Royal blue", "Dark blue", "Blue", "Cyan", "Green", "Lime", "PC Amber", "Red", "Deep red"))

mapped_settings <- mapped_settings %>% 
  left_join(peak_nm_df, by = "primary")
mapped_settings
```

# Calculate coverage in log10 units

```{r calculate-log-coverage}
irradiance_levels_log10 <- log10(mapped_settings$irradiance_theoretical)
coverage <- max(irradiance_levels_log10) - min(irradiance_levels_log10)
coverage
```

# Check settings

Here we did calibration measurements to check whether the calculated settings are as expected in terms of output

```{r check-target-settings, fig.width=5, fig.height=3}
# calculated settings and theoretical irradiance, illuminance and mel EDI 
theoretical <- mapped_settings

merge_theoretical_measured <- function(theoretical_data, measured_data_filename) {
  
  data_measured <- prepare_outcomes(measured_data_filename) %>% 
    mutate(measurement_num = as.numeric(measurement)) %>% 
    rename(
      irradiance_measured = `Ee [W/(sqm)]`,
      illuminance_measured = `Illuminance (lx)`,
      MEDI_measured = `Melanopic EDI (lx)`,
      REDI_measured = `Rhodopic EDI (lx)`
    )
  
  data_measured <- data_measured %>%
    mutate(
      repetition = ceiling(measurement_num / (6 * 4)),
      within_rep = measurement_num - (repetition - 1) * 24,
      primary_num = case_when(
        repetition <= 3 ~ ceiling(within_rep / 4) + 1,
        repetition > 3  ~ 8 - ceiling(within_rep / 4)
      ),
      lab_visit_num = ((within_rep - 1) %% 4) + 1,
      lab_visit = case_when(
        repetition <= 3 ~ as.character(lab_visit_num),
        repetition > 3  ~ as.character(5 - lab_visit_num)
      ),
      primary = factor(primary_num, levels = as.character(2:7)),
      lab_visit = factor(lab_visit, levels = as.character(1:4))
    )

  # plot <- data_measured %>% 
  #   ggplot(aes(x = primary, y = irradiance_measured, color = lab_visit)) +
  #   geom_point() +
  #   scale_y_log10(breaks = breaks, minor_breaks = minor_breaks) +
  #   scale_colour_manual(values = colorRampPalette(c("gold", "red3"))(4)) +
  #   labs(x = "Primary",
  #        colour = "Light intensity") +
  #   theme_minimal() +
  #   plot_formatting
  # 
  # print(plot)
  
  data_means <- data_measured %>% 
    group_by(primary, lab_visit) %>% 
    summarise(
      # Arithmetic means
      irradiance_measured_mean = mean(irradiance_measured, na.rm = TRUE),
      illuminance_measured_mean = mean(illuminance_measured, na.rm = TRUE),
      MEDI_measured_mean = mean(MEDI_measured, na.rm = TRUE),
      REDI_measured_mean = mean(REDI_measured, na.rm = TRUE),
  
      # Geometric means
      irradiance_measured_gmean = exp(mean(log(irradiance_measured), na.rm = TRUE)),
      illuminance_measured_gmean = exp(mean(log(illuminance_measured), na.rm = TRUE)),
      MEDI_measured_gmean = exp(mean(log(MEDI_measured), na.rm = TRUE)),
      REDI_measured_gmean = exp(mean(log(REDI_measured), na.rm = TRUE)),
  
      # Geometric SDs
      irradiance_measured_gsd = exp(sd(log(irradiance_measured), na.rm = TRUE)),
      illuminance_measured_gsd = exp(sd(log(illuminance_measured), na.rm = TRUE)),
      MEDI_measured_gsd = exp(sd(log(MEDI_measured), na.rm = TRUE)),
      REDI_measured_gsd = exp(sd(log(REDI_measured), na.rm = TRUE)),
  
      .groups = "drop"
    )
  
  data_merged <- data_means %>% 
    left_join(theoretical_data, by = c("primary", "lab_visit"))
  
  # Ratios
  data_merged <- data_merged %>% 
    mutate(
      ratio_mel_pho = MEDI_measured_gmean / illuminance_measured_gmean,
      ratio_rho_pho = REDI_measured_gmean / illuminance_measured_gmean,
      ratio_mel_rho = MEDI_measured_gmean / REDI_measured_gmean
    )
  
  return(data_merged)
}

data_measured_1 <- merge_theoretical_measured(
  theoretical_data = theoretical, 
  measured_data_filename = "2a_target_settings")
```

# Visualise differences between calculated and measured 

```{r compare-calculated-measured}
plot_diff_measured <- function(df, save = FALSE) {
  
  col_vals <- colorRampPalette(c("gold", "red3"))(4)
  
  # Irradiance
  plot_ee <- df %>% 
    ggplot(aes(
      x = irradiance_theoretical,
      y = irradiance_measured_gmean,
      colour = lab_visit
    )) +
    geom_errorbar(aes(
      ymin = irradiance_measured_gmean / irradiance_measured_gsd,
      ymax = irradiance_measured_gmean * irradiance_measured_gsd
    ),
    width = 0.2,
    alpha = 0.7
    ) +
    geom_point(size = 2) +
    geom_abline(intercept = 0, slope = 1, color = "black") +
    facet_wrap(~ primary) +
    scale_y_log10(breaks = breaks, minor_breaks = minor_breaks) +
    scale_x_log10(breaks = breaks, minor_breaks = minor_breaks) +
    scale_colour_manual(values = col_vals) +
    labs(
      x = "Theoretical",
      y = "Measured",
      title = "Irradiance [W/sqm] (380-780 nm)",
      colour = "Light intensity"
    ) +
    theme_minimal() +
    plot_formatting
  
  # Illuminance
  plot_ev <- df %>% 
    ggplot(aes(
      x = illuminance_theoretical,
      y = illuminance_measured_gmean,
      colour = lab_visit
    )) +
    geom_errorbar(aes(
      ymin = illuminance_measured_gmean / illuminance_measured_gsd,
      ymax = illuminance_measured_gmean * illuminance_measured_gsd
    ),
    width = 0.2,
    alpha = 0.7
    ) +
    geom_point(size = 2) +
    geom_abline(intercept = 0, slope = 1, color = "black") +
    facet_wrap(~ primary) +
    scale_y_log10(breaks = breaks, minor_breaks = minor_breaks) +
    scale_x_log10(breaks = breaks, minor_breaks = minor_breaks) +
    scale_colour_manual(values = col_vals) +
    labs(
      x = "Theoretical",
      y = "Measured",
      title = "Illuminance [lx]",
      colour = "Light intensity"
    ) +
    theme_minimal() +
    plot_formatting
  
  # Melanopic EDI
  plot_medi <- df %>% 
    ggplot(aes(
      x = MEDI_theoretical,
      y = MEDI_measured_gmean,
      colour = lab_visit
    )) +
    geom_errorbar(aes(
      ymin = MEDI_measured_gmean / MEDI_measured_gsd,
      ymax = MEDI_measured_gmean * MEDI_measured_gsd
    ),
    width = 0.2,
    alpha = 0.7
    ) +
    geom_point(size = 2) +
    geom_abline(intercept = 0, slope = 1, color = "black") +
    facet_wrap(~ primary) +
    scale_y_log10(breaks = breaks, minor_breaks = minor_breaks) +
    scale_x_log10(breaks = breaks, minor_breaks = minor_breaks) +
    scale_colour_manual(values = col_vals) +
    labs(
      x = "Theoretical",
      y = "Measured",
      title = "Melanopic EDI [lx]",
      colour = "Light intensity"
    ) +
    theme_minimal() +
    plot_formatting
  
  # Rhodopic EDI
  plot_redi <- df %>% 
    ggplot(aes(
      x = REDI_theoretical,
      y = REDI_measured_gmean,
      colour = lab_visit
    )) +
    geom_errorbar(aes(
      ymin = REDI_measured_gmean / REDI_measured_gsd,
      ymax = REDI_measured_gmean * REDI_measured_gsd
    ),
    width = 0.2,
    alpha = 0.7
    ) +
    geom_point(size = 2) +
    geom_abline(intercept = 0, slope = 1, color = "black") +
    facet_wrap(~ primary) +
    scale_y_log10(breaks = breaks, minor_breaks = minor_breaks) +
    scale_x_log10(breaks = breaks, minor_breaks = minor_breaks) +
    scale_colour_manual(values = col_vals) +
    labs(
      x = "Theoretical",
      y = "Measured",
      title = "Rhodopic EDI [lx]",
      colour = "Light intensity"
    ) +
    theme_minimal() +
    plot_formatting
  
  print(plot_ee)
  print(plot_ev)
  print(plot_medi)
  print(plot_redi)
  
  if (save) {
    formatted_save(here("outputs", "figures", "calibration_ee_accuracy.jpg"), plot_ee)
    formatted_save(here("outputs", "figures", "calibration_ev_accuracy.jpg"), plot_ev)
    formatted_save(here("outputs", "figures", "calibration_medi_accuracy.jpg"), plot_medi)
    formatted_save(here("outputs", "figures", "calibration_redi_accuracy.jpg"), plot_redi)
  }
}
``` 

```{r plot-calculated-measured-differences, fig.width=5, fig.height=3}
plot_diff_measured(data_measured_1)
```

# Visualise differences in measured metrics between primaries

```{r plot-primary-metric-differences}
plot_diff_metrics <- function(df, save = FALSE) {
  
  # ----------------------------
  # Helpers
  # ----------------------------
  metric_palette <- colorRampPalette(c("gold", "red3"))(4)
  
  add_gsd_bounds <- function(data, gmean_var, gsd_var) {
    data %>%
      mutate(
        y = .data[[gmean_var]],
        ymin = .data[[gmean_var]] / .data[[gsd_var]],
        ymax = .data[[gmean_var]] * .data[[gsd_var]]
      )
  }
  
  make_metric_plot <- function(data, gmean_var, gsd_var, title, y_lab) {
    
    plot_data <- add_gsd_bounds(data, gmean_var, gsd_var)
    
    ggplot(
      plot_data,
      aes(
        x = primary,
        y = y,
        colour = lab_visit,
        group = lab_visit
      )
    ) +
      geom_errorbar(
        aes(ymin = ymin, ymax = ymax),
        width = 0.2,
        alpha = 0.7
      ) +
      geom_point() +
      geom_line(alpha = 0.3) +
      scale_y_log10(breaks = breaks, minor_breaks = minor_breaks) +
      scale_colour_manual(values = metric_palette) +
      labs(
        title = title,
        y = y_lab,
        x = "Primary",
        colour = "Light intensity"
      ) +
      theme_minimal() +
      plot_formatting
  }
  
  make_comparison_plot <- function(data, var_1, gsd_1, var_2, gsd_2, label_1, label_2, title) {
    
    plot_data <- bind_rows(
      data %>%
        transmute(
          primary,
          lab_visit,
          metric_type = label_1,
          value = .data[[var_1]],
          ymin = .data[[var_1]] / .data[[gsd_1]],
          ymax = .data[[var_1]] * .data[[gsd_1]]
        ),
      data %>%
        transmute(
          primary,
          lab_visit,
          metric_type = label_2,
          value = .data[[var_2]],
          ymin = .data[[var_2]] / .data[[gsd_2]],
          ymax = .data[[var_2]] * .data[[gsd_2]]
        )
    )
    
    ggplot(
      plot_data,
      aes(
        x = primary,
        y = value,
        ymin = ymin,
        ymax = ymax,
        shape = metric_type,
        # linetype = metric_type,
        colour = lab_visit,
        group = interaction(lab_visit, metric_type)
      )
    ) +
      geom_errorbar(
        width = 0.2,
        alpha = 0.7
      ) +
      geom_point() +
      geom_line(alpha = 0.3) +
      scale_y_log10(breaks = breaks, minor_breaks = minor_breaks) +
      scale_colour_manual(values = metric_palette) +
      scale_shape_manual(
        values = c(
          "Illuminance"    = 1,   # empty circle
          "Melanopic EDI" = 17,   # filled triangle
          "Rhodopic EDI"  = 15    # filled square
        ),
        breaks = c("Illuminance", "Melanopic EDI", "Rhodopic EDI")
      ) +
      labs(
        title = title,
        y = "[lx]",
        x = "Primary",
        colour = "Light intensity",
        # linetype = "Metric",
        shape = "Metric"
      ) +
      theme_minimal() +
      theme(legend.key.height = unit(0.4, "cm")) +
      plot_formatting
  }
  
  make_ratio_plot <- function(data, y_var, title) {
    ggplot(
      data,
      aes(
        x = primary,
        y = .data[[y_var]],
        colour = lab_visit
      )
    ) +
      geom_hline(yintercept = 1, colour = "darkred") +
      geom_point() +
      scale_y_log10(breaks = c(0.1, 0.3, 1, 3, 10), minor_breaks = minor_breaks) +
      scale_colour_manual(values = metric_palette) +
      labs(
        title = title,
        y = "Ratio",
        x = "Primary",
        colour = "Light intensity"
      ) +
      coord_cartesian(ylim = c(0.1, 10)) +
      theme_minimal() +
      plot_formatting
  }
  
  # ----------------------------
  # Single-metric plots
  # ----------------------------
  plot_irradiance <- make_metric_plot(
    data     = df,
    gmean_var = "irradiance_measured_gmean",
    gsd_var   = "irradiance_measured_gsd",
    title    = "Irradiance",
    y_lab    = "[W/sqm] (380-780nm)"
  )
  
  plot_illuminance <- make_metric_plot(
    data      = df,
    gmean_var = "illuminance_measured_gmean",
    gsd_var   = "illuminance_measured_gsd",
    title     = "Illuminance",
    y_lab     = "[lx]"
  )
  
  plot_melanopic_edi <- make_metric_plot(
    data      = df,
    gmean_var = "MEDI_measured_gmean",
    gsd_var   = "MEDI_measured_gsd",
    title     = "Melanopic EDI",
    y_lab     = "[lx]"
  )
  
  plot_rhodopic_edi <- make_metric_plot(
    data      = df,
    gmean_var = "REDI_measured_gmean",
    gsd_var   = "REDI_measured_gsd",
    title     = "Rhodopic EDI",
    y_lab     = "[lx]"
  )
  
  # ----------------------------
  # Two-metric comparison plots
  # ----------------------------
  plot_melanopic_vs_photopic <- make_comparison_plot(
    data    = df,
    var_1   = "MEDI_measured_gmean",
    gsd_1   = "MEDI_measured_gsd",
    var_2   = "illuminance_measured_gmean",
    gsd_2   = "illuminance_measured_gsd",
    label_1 = "Melanopic EDI",
    label_2 = "Illuminance",
    title   = "Melanopic vs Photopic"
  )
  
  plot_rhodopic_vs_photopic <- make_comparison_plot(
    data    = df,
    var_1   = "REDI_measured_gmean",
    gsd_1   = "REDI_measured_gsd",
    var_2   = "illuminance_measured_gmean",
    gsd_2   = "illuminance_measured_gsd",
    label_1 = "Rhodopic EDI",
    label_2 = "Illuminance",
    title   = "Rhodopic vs Photopic"
  )
  
  plot_melanopic_vs_rhodopic <- make_comparison_plot(
    data    = df,
    var_1   = "MEDI_measured_gmean",
    gsd_1   = "MEDI_measured_gsd",
    var_2   = "REDI_measured_gmean",
    gsd_2   = "REDI_measured_gsd",
    label_1 = "Melanopic EDI",
    label_2 = "Rhodopic EDI",
    title   = "Melanopic vs Rhodopic"
  )
  
  # ----------------------------
  # Ratio plots
  # ----------------------------
  plot_ratio_melanopic_photopic <- make_ratio_plot(
    data  = df,
    y_var = "ratio_mel_pho",
    title = "Melanopic vs Photopic"
  )
  
  plot_ratio_rhodopic_photopic <- make_ratio_plot(
    data  = df,
    y_var = "ratio_rho_pho",
    title = "Rhodopic vs Photopic"
  )
  
  plot_ratio_melanopic_rhodopic <- make_ratio_plot(
    data  = df,
    y_var = "ratio_mel_rho",
    title = "Melanopic vs Rhodopic"
  )
  
  # ----------------------------
  # Print plots
  # ----------------------------
  print(plot_irradiance)
  print(plot_illuminance)
  print(plot_melanopic_edi)
  print(plot_rhodopic_edi)
  print(plot_melanopic_vs_photopic)
  print(plot_rhodopic_vs_photopic)
  print(plot_melanopic_vs_rhodopic)
  print(plot_ratio_melanopic_photopic)
  print(plot_ratio_rhodopic_photopic)
  print(plot_ratio_melanopic_rhodopic)
  
  # ----------------------------
  # Save plots
  # ----------------------------
  if (save) {
    save_rds_safe(
      plot_irradiance,
      here("outputs", "figures", "rds", "calibration_irradiance.rds")
    )
    
    save_rds_safe(
      plot_illuminance,
      here("outputs", "figures", "rds", "calibration_illuminance.rds")
    )
    
    save_rds_safe(
      plot_melanopic_edi,
      here("outputs", "figures", "rds", "calibration_melanopic_edi.rds")
    )
    
    save_rds_safe(
      plot_rhodopic_edi,
      here("outputs", "figures", "rds", "calibration_rhodopic_edi.rds")
    )
    
    save_rds_safe(
      plot_melanopic_vs_photopic,
      here("outputs", "figures", "rds", "calibration_melanopic_vs_photopic.rds")
    )
    
    save_rds_safe(
      plot_rhodopic_vs_photopic,
      here("outputs", "figures", "rds", "calibration_rhodopic_vs_photopic.rds")
    )
    
    save_rds_safe(
      plot_melanopic_vs_rhodopic,
      here("outputs", "figures", "rds", "calibration_melanopic_vs_rhodopic.rds")
    )
    
    save_rds_safe(
      plot_ratio_melanopic_photopic,
      here("outputs", "figures", "rds", "calibration_ratio_melanopic_photopic.rds")
    )
    
    save_rds_safe(
      plot_ratio_rhodopic_photopic,
      here("outputs", "figures", "rds", "calibration_ratio_rhodopic_photopic.rds")
    )
    
    save_rds_safe(
      plot_ratio_melanopic_rhodopic,
      here("outputs", "figures", "rds", "calibration_ratio_melanopic_rhodopic.rds")
    )
  }
  
  invisible(list(
    plot_irradiance = plot_irradiance,
    plot_illuminance = plot_illuminance,
    plot_melanopic_edi = plot_melanopic_edi,
    plot_rhodopic_edi = plot_rhodopic_edi,
    plot_melanopic_vs_photopic = plot_melanopic_vs_photopic,
    plot_rhodopic_vs_photopic = plot_rhodopic_vs_photopic,
    plot_melanopic_vs_rhodopic = plot_melanopic_vs_rhodopic,
    plot_ratio_melanopic_photopic = plot_ratio_melanopic_photopic,
    plot_ratio_rhodopic_photopic = plot_ratio_rhodopic_photopic,
    plot_ratio_melanopic_rhodopic = plot_ratio_melanopic_rhodopic
  ))
}
```

```{r create-final-calibration-plot, fig.width=5, fig.height=3}
plot_diff_metrics(data_measured_1)
```

We see that for the second lab visit, the irradiance is not matched…
So, we manually correct it and check with new spectral measurements

```{r check-corrected-settings, fig.width=5, fig.height=3}
data_measured_2 <- merge_theoretical_measured(
  theoretical_data = theoretical, 
  measured_data_filename = "2b_target_settings_corrected"
) %>% 
  rename(setting_theoretical = setting) %>% 
  # Update the settings from the old DF that have been manually adjusted
  mutate(
    setting_corrected = setting_theoretical,
    setting_corrected = case_when(lab_visit == 2 & primary == 2 ~ 128,
                                  lab_visit == 2 & primary == 3 ~ 142,
                                  lab_visit == 2 & primary == 4 ~ 156,
                                  lab_visit == 2 & primary == 5 ~ 166,
                                  lab_visit == 2 & primary == 6 ~ 174,
                                  lab_visit == 2 & primary == 7 ~ 115,
                                  TRUE ~ setting_corrected)
  )

plot_diff_measured(data_measured_2)
plot_diff_metrics(data_measured_2, save = T)
```

There is still a discrepancy between theoretical and measured metrics.
Main thing is that the irradiance is constant across primaries, and well log-spaced between lab visits.

# Create and save final plot

```{r import-final-plots}
# import plots
plot_irradiance <- readRDS(here("outputs/figures/rds", "calibration_irradiance.rds"))
plot_ratio_mel_pho <- readRDS(here("outputs/figures/rds", "calibration_melanopic_vs_photopic.rds"))
plot_ratio_rho_pho <- readRDS(here("outputs/figures/rds", "calibration_rhodopic_vs_photopic.rds"))
plot_ratio_mel_rho <- readRDS(here("outputs/figures/rds", "calibration_melanopic_vs_rhodopic.rds"))
```

```{r assemble-final-calibration-plot}
plot_irradiance_2 <- plot_irradiance + 
  labs(title = expression(bold("A ") * "Irradiance")) + 
  # guides(colour = "none") +
  plot_formatting

plot_ratio_mel_pho_2 <- plot_ratio_mel_pho +
  labs(title = expression(bold("B ") * "Melanopic vs Photopic")) + 
  guides(colour = "none") +
  plot_formatting

plot_ratio_rho_pho_2 <- plot_ratio_rho_pho +
  labs(title = expression(bold("C ") * "Rhodopic vs Photopic")) + 
  guides(colour = "none") +
  plot_formatting

plot_ratio_mel_rho_2 <- plot_ratio_mel_rho +
  labs(title = expression(bold("D ") * "Melanopic vs Rhodopic")) + 
  guides(colour = "none") +
  plot_formatting

plot_all <- plot_irradiance_2 + plot_ratio_mel_pho_2 + plot_ratio_rho_pho_2 + plot_ratio_mel_rho_2
plot_all
```

```{r save-final-calibration-plot}
formatted_save(here("outputs/figures", "calibration_light_metrics.png"), plot_all)
save_rds_safe(plot_all, here("outputs", "figures", "rds", glue("calibration_light_metrics.rds")))
```

# Generate a design matrix

```{r generate-design-matrix}
# -----------------------------
# Generate DeBruijn sequences
# -----------------------------

# Run the standalone DeBruijn sequence-generation script.
# This creates outputs/debruijn_sequences.csv
source(
  here("R", "generate_debruijn_sequence.R")
)

debruijn_sequences <- readr::read_csv(
  here("outputs", "debruijn_sequences.csv"),
  show_col_types = FALSE
)

# -----------------------------
# Keep only the number of sequences needed
# -----------------------------

n_participants <- 100

lab_visits <- sort(unique(data_measured_2$lab_visit))

n_sessions <- n_participants * length(lab_visits)

if (dplyr::n_distinct(debruijn_sequences$sequence_id) < n_sessions) {
  stop(
    "Not enough DeBruijn sequences. Need ",
    n_sessions,
    " unique sequences, but found only ",
    dplyr::n_distinct(debruijn_sequences$sequence_id),
    ". Increase n_sequences in generate_debruijn_sequence.R.",
    call. = FALSE
  )
}

participant_sequence_map <- tidyr::crossing(
  participant_id = factor(seq_len(n_participants)),
  lab_visit = lab_visits
) %>%
  dplyr::arrange(participant_id, lab_visit) %>%
  dplyr::mutate(
    sequence_id = dplyr::row_number()
  )

debruijn_sequences_final <- debruijn_sequences %>%
  dplyr::inner_join(
    participant_sequence_map,
    by = "sequence_id"
  )

# -----------------------------
# Create one-row-per-condition-per-visit lookup table
# -----------------------------

condition_lookup <- data_measured_2 %>% 
  dplyr::select(
    lab_visit,
    primary,
    peak_nm,
    color,
    setting_corrected,
    irradiance_measured_gmean,
    illuminance_measured_gmean,
    MEDI_measured_gmean,
    REDI_measured_gmean,
    ratio_mel_pho,
    ratio_rho_pho,
    ratio_mel_rho
  ) %>% 
  dplyr::rename(
    setting = setting_corrected,
    irradiance = irradiance_measured_gmean,
    illuminance = illuminance_measured_gmean,
    MEDI = MEDI_measured_gmean,
    REDI = REDI_measured_gmean
  ) %>%
  dplyr::group_by(lab_visit) %>%
  dplyr::arrange(primary, .by_group = TRUE) %>%
  dplyr::mutate(
    assigned_condition = dplyr::row_number()
  ) %>%
  dplyr::ungroup()

# Check that each lab visit has 6 condition rows
condition_counts <- condition_lookup %>%
  dplyr::count(lab_visit, name = "n_conditions")

if (any(condition_counts$n_conditions != 6)) {
  stop(
    "Expected 6 condition rows per lab_visit, but found:\n",
    paste(
      paste0(condition_counts$lab_visit, ": ", condition_counts$n_conditions),
      collapse = "\n"
    ),
    call. = FALSE
  )
}

# -----------------------------
# Join trial sequences to measured condition information
# -----------------------------

design_matrix <- debruijn_sequences_final %>%
  dplyr::left_join(
    condition_lookup,
    by = c("lab_visit", "assigned_condition")
  ) %>%
  dplyr::arrange(participant_id, lab_visit, trial)

# Check that every trial received condition information
if (any(is.na(design_matrix$setting))) {
  stop(
    "Some DeBruijn trials did not match a condition in condition_lookup.",
    call. = FALSE
  )
}

# Check expected final size
expected_n_rows <- n_participants * length(lab_visits) * 6^2

if (nrow(design_matrix) != expected_n_rows) {
  warning(
    "Expected ", expected_n_rows, " rows, but got ", nrow(design_matrix), "."
  )
}

# Optional sanity check:
# each participant × lab_visit should have 36 trials
session_counts <- design_matrix %>%
  dplyr::count(participant_id, lab_visit, name = "n_trials")

if (any(session_counts$n_trials != 6^2)) {
  stop(
    "Some participant × lab_visit sessions do not have 36 trials.",
    call. = FALSE
  )
}

# -----------------------------
# Save
# -----------------------------

readr::write_csv(
  design_matrix,
  here("outputs", "design_matrix.csv")
)
```

```{r session-info}
sessionInfo()
```

```{r render-time-2, echo=FALSE}
render_time <- Sys.time() - .render_start
cat("Render time:", format(render_time))
```
