---
title: "EMG modelling"
author: "Johannes Zauner, Lucien Bickerstaff"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
---

```{r render-time-1, include=FALSE}
# Record how long it took to render the file
.render_start <- Sys.time()
```

```{r, include=FALSE}
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE
)
```

# Setup

```{r setup}
# Seed for reproducibility
set.seed(2025)

# Libraries
library(here)
library(tidyverse)
library(lme4)
library(glmmTMB)
library(performance)
library(visreg)
library(emmeans)
library(gtsummary)
library(ggeffects)
library(patchwork)
library(LightLogR)
library(scales)

source(here("R", "helpers.R"))
source(here("R", "plot_settings.R"))
```

# Read EMG data

```{r}
# ------------------------------------------------------------------------------
# Prepare trial-level EMG data for modelling
# ------------------------------------------------------------------------------

# Import trial-level EMG outcomes
emg_data <- readRDS(
  here::here("data", "main_data", "emg_outcomes_trial.rds")
)

# Keep the EMG channel of interest and remove calibration variables that will be
# replaced with Luox-derived calibration estimates.
emg_trial_data <- emg_data %>%
  dplyr::filter(channel == 2) %>%
  dplyr::select(
    -melanopic_edi,
    -illuminance,
    -emg_trace
  ) %>%
  dplyr::mutate(
    dplyr::across(
      c(participant_id, condition),
      factor
    )
  )

# Import Luox-derived calibration values for each light condition.
#
# The calibration file is stored in a wide format where rows are measurement
# types and columns are light conditions. The data are reshaped so that each row
# corresponds to one experimental condition.
calibration_data <- readr::read_csv(
  here::here(
    "data",
    "light_source_calibration",
    "3_luox_outputs",
    "3_pilot_settings_luox.csv"
  )
) %>%
  dplyr::mutate(
    dplyr::across(
      -Condition,
      as.numeric
    )
  ) %>%
  tidyr::pivot_longer(
    cols = -Condition,
    names_to = "measurement",
    values_to = "value"
  ) %>%
  tidyr::pivot_wider(
    names_from = Condition,
    values_from = value
  ) %>%
  dplyr::mutate(
    condition = factor(dplyr::row_number() - 1)
  ) %>%
  dplyr::select(
    condition,
    illuminance   = `Illuminance (lx)`,
    rhodopic_edi  = `Rhodopic EDI (lx)`,
    melanopic_edi = `Melanopic EDI (lx)`
  ) %>%
  dplyr::mutate(
    # Treat very small measured values as the nominal 0-lx condition.
    dplyr::across(
      c(illuminance, rhodopic_edi, melanopic_edi),
      ~ dplyr::if_else(.x < 1, 0, .x)
    ),

    # Rounded value used for grouping/visualisation, not modelling.
    melanopic_edi_round = round(melanopic_edi)
  )

# Add calibrated light-source information to the trial-level EMG data.
data <- emg_trial_data %>%
  dplyr::left_join(
    calibration_data,
    by = "condition"
  )

# Create the modelling dataset.
#
# Rows with non-positive EMG values are excluded because log10(emg_raw) is only
# defined for strictly positive values. For melanopic EDI, the 0-lx condition is
# retained by using log10(pmax(melanopic_edi, 1)); this maps all values below 1 lx
# to 0 on the log10 scale.
model_data <- data %>%
  dplyr::filter(
    emg_raw > 0
  ) %>%
  dplyr::mutate(
    lights_on = melanopic_edi != 0,

    log_emg_raw       = log10(emg_raw),
    log_melanopic_edi = log10(pmax(melanopic_edi, 1)),
    log_illuminance   = log10(pmax(illuminance, 1)),
    log_rhodopic_edi  = log10(pmax(rhodopic_edi, 1)),
    log_baseline      = log10(baseline_value)
  )

# Check how many rows were excluded by the log-transform requirement.
excluded_rows <- nrow(data) - nrow(model_data)

excluded_rows
```

`excluded_rows` records how many rows were removed because one or more variables required by the log-transformed model were missing or non-positive.

## Visualise

```{r}
#| warning: false
#| message: false
#| fig-width: 12
#| fig-height: 12

Plot_unscaled <-
  data |> 
  ggplot(aes(y = emg_raw, x = melanopic_edi_round)) +
  geom_violin(aes(group = melanopic_edi_round), scale = "width") +
  stat_summary(aes(group = melanopic_edi_round), fun = mean) +
  stat_summary(aes(group = melanopic_edi_round), fun = median, col = "red", size = 0.5) +
  labs(
    title = "Violin plots of response against stimulus with mean (black) and median (red) response\nblue: linear model with 0lx, green: linear model without 0lx"
  ) +
  geom_smooth(method = "lm") +
  geom_smooth(
    aes(group = melanopic_edi_round == 0),
    col = "green4",
    fill = "green4",
    method = "lm"
  )

Plot_scaled <- 
  Plot_unscaled +
  scale_x_continuous(trans = "symlog", breaks = c(0, 10^(0:5))) +
  labs(title = NULL)

Plot_unscaled_log10 <-
  Plot_unscaled +
  scale_y_log10() +
  labs(title = NULL, subtitle = "Y scaled log10")
  
Plot_scaled_log10 <- 
  Plot_scaled +
  scale_y_log10()

(Plot_unscaled + Plot_scaled) /
  (Plot_unscaled_log10 + Plot_scaled_log10)
```

# Model

## Helpers

```{r}
fit_glmmTMB <- function(formula, data) {
  data_name <- substitute(data)

  model <- glmmTMB::glmmTMB(
    formula = formula,
    data = data
  )

  # Make downstream functions like visreg() able to find the original data
  model$call$data <- data_name

  model
}

check_glmmTMB <- function(model, save_path = NULL) {
  print(summary(model))

  cat("\nConvergence check:\n")
  print(performance::check_convergence(model))

  cat("\nSingularity check:\n")
  print(performance::check_singularity(model))

  diagnostic_plot <- performance::check_model(model)
  # print(diagnostic_plot)
  
  print(performance::r2(model))
  print(performance::icc(model))
  print(performance::check_collinearity(model))
  
  if (!is.null(save_path)) {
    ggsave_safe(
      save_path, 
      plot(diagnostic_plot),
      width = 9,
      height = 9,
      dpi = 300)
  }

  invisible(model)
}
```

## Random effects for melanopic EDI?

```{r}
#| fig-width: 8
#| fig-height: 12

form_m0 <- log_emg_raw ~
  log_melanopic_edi +
  lights_on

m0 <- fit_glmmTMB(form_m0, model_data)

check_glmmTMB(m0)
```

```{r}
#| fig-width: 8
#| fig-height: 12

form_m1 <- log_emg_raw ~
  log_melanopic_edi +
  lights_on +
  (log_melanopic_edi | participant_id)

m1 <- fit_glmmTMB(form_m1, model_data)

check_glmmTMB(m1)
```

```{r}
#| fig-width: 8
#| fig-height: 12

form_m2 <- log_emg_raw ~
  log_melanopic_edi +
  lights_on +
  (1 | participant_id)

m2 <- fit_glmmTMB(form_m2, model_data)

check_glmmTMB(m2)
```

```{r}
AIC(m0, m1, m2)
```

`m1` is the preferred model (lower AIC).
Keeping random slopes is warranted here.

## Include baseline?

```{r}
#| fig-width: 8
#| fig-height: 12

form_m3 <- log_emg_raw ~
  log_melanopic_edi +
  lights_on +
  log_baseline +
  (log_melanopic_edi | participant_id)

m3 <- fit_glmmTMB(form_m3, model_data)

check_glmmTMB(m3)
```

```{r}
AIC(m1, m3)
```

`m3` is the preferred model.
Including baseline is warranted here.

## Include trial number?

```{r}
#| fig-width: 8
#| fig-height: 12

form_m4 <- log_emg_raw ~
  log_melanopic_edi +
  lights_on +
  log_baseline +
  trial +
  (log_melanopic_edi | participant_id)

m4 <- fit_glmmTMB(form_m4, model_data)

check_glmmTMB(m4, save_path = here::here("outputs", "models", "emg_lmm_mel_diag.png"))
```

```{r}
AIC(m3, m4)
```

`m4` is the better fit.
Including trial number is warranted.

```{r}
#| fig-width: 8
#| fig-height: 6

visreg(m4, xvar = "log_melanopic_edi", jitter = TRUE)
visreg(m4, xvar = "lights_on", jitter = TRUE)
```

# Why not a model with all predictors?

```{r}
#| fig-width: 8
#| fig-height: 12

form_m5 <- log_emg_raw ~
  log_melanopic_edi +
  log_illuminance +
  log_rhodopic_edi +
  lights_on +
  log_baseline +
  trial +
  (log_melanopic_edi | participant_id) +
  (log_illuminance | participant_id) +
  (log_rhodopic_edi | participant_id)

m5 <- fit_glmmTMB(form_m5, model_data)

check_glmmTMB(m5)
```

A model with all predictors suffers from collinearity issues, because the
melanopic and rhodopic curves are very close to each other and the model
cannot decide between the two…

# Predictions against actual data

```{r}
#| warning: false
#| message: false
#| fig-width: 8
#| fig-height: 5

# ------------------------------------------------------------------------------
# Helper functions
# ------------------------------------------------------------------------------

make_prediction_grid <- function(data, n = 1000) {
  # Reference values for fixed-effect predictions.
  #
  # These hold non-focal predictors constant while melanopic EDI varies.
  baseline_reference <- median(data$baseline_value, na.rm = TRUE)
  trial_reference    <- median(data$trial, na.rm = TRUE)

  max_melanopic_edi <- max(data$melanopic_edi, na.rm = TRUE)

  pred_df <- tibble::tibble(
    # Include 0 manually so the lights-off condition is represented.
    # The continuous grid starts at 0.01 because log10(0) is undefined.
    melanopic_edi = c(
      0,
      10^seq(
        from = -2,
        to = log10(max_melanopic_edi),
        length.out = n
      )
    ),
    log_baseline = log10(baseline_reference),
    trial = trial_reference
  ) |>
    dplyr::mutate(
      lights_on = melanopic_edi != 0,

      # Must match the transformation used in the model.
      log_melanopic_edi = log10(pmax(melanopic_edi, 1))
    )

  attr(pred_df, "baseline_reference") <- baseline_reference
  attr(pred_df, "trial_reference")    <- trial_reference

  pred_df
}


add_fixed_effect_predictions <- function(pred_df, model) {
  # Predictions are fixed-effect only because re.form = NA.
  #
  # If the outcome model is log10(emg_raw), then type = "response" returns
  # predictions on the log10 EMG scale. We back-transform with 10^x.
  pred_df |>
    dplyr::mutate(
      pred_log_emg = predict(
        model,
        newdata = dplyr::pick(dplyr::everything()),
        re.form = NA,
        type = "response"
      ),
      pred_emg = 10^pred_log_emg
    )
}


make_pseudolog_breaks <- function(thres = 2, max_power = 5) {
  list(
    breaks = c(
      0,
      1,
      2,
      10^(1:max_power)
    ),
    minor_breaks = c(
      0,
      seq(1, thres, by = 0.5),
      1:9 * 10^rep(1:max_power, each = 9)
    )
  )
}


plot_observed_vs_predicted <- function(data,
                                       pred_df,
                                       model_name = "m4",
                                       thres = 2,
                                       log_y = FALSE) {
  # Rounded EDI is used only to group observed points and violins.
  # Do not use this rounded variable for modelling or prediction.
  data_plot <- data |>
    dplyr::mutate(
      melanopic_edi_plot = round(melanopic_edi)
    )

  x_breaks <- make_pseudolog_breaks(thres = thres)

  baseline_reference <- attr(pred_df, "baseline_reference")
  trial_reference    <- attr(pred_df, "trial_reference")

  p <- data_plot |>
    ggplot2::ggplot(
      ggplot2::aes(
        x = melanopic_edi_plot,
        y = emg_raw
      )
    ) +
    ggplot2::geom_point(
      alpha = 0.35,
      size = 1.5
    ) +
    ggplot2::geom_violin(
      ggplot2::aes(group = melanopic_edi_plot),
      scale = "width",
      alpha = 0.25
    ) +
    ggplot2::stat_summary(
      ggplot2::aes(group = melanopic_edi_plot),
      fun = mean,
      geom = "point",
      colour = "black",
      size = 2
    ) +
    ggplot2::stat_summary(
      ggplot2::aes(group = melanopic_edi_plot),
      fun = median,
      geom = "point",
      colour = "red",
      size = 2
    ) +
    ggplot2::geom_line(
      data = pred_df,
      ggplot2::aes(
        x = melanopic_edi,
        y = pred_emg
      ),
      inherit.aes = FALSE,
      linewidth = 1,
      colour = "blue",
      alpha = 0.7
    ) +
    ggplot2::scale_x_continuous(
      trans = scales::pseudo_log_trans(
        base = 10,
        sigma = thres
      ),
      breaks = x_breaks$breaks,
      minor_breaks = x_breaks$minor_breaks
    ) +
    ggplot2::labs(
      x = "Melanopic EDI",
      y = "Raw EMG",
      title = "Observed data vs model prediction",
      subtitle = paste0(
        "Blue line: fixed-effect prediction from ", model_name, "; ",
        "baseline EMG held at median = ", signif(baseline_reference, 3), "; ",
        "trial held at median = ", signif(trial_reference, 3), "; ",
        "lights_on defined as melanopic_edi != 0"
      )
    ) +
    ggplot2::theme_minimal()

  if (log_y) {
    p <- p +
      ggplot2::scale_y_log10() +
      ggplot2::labs(
        title = sprintf(
          "Observed data vs model prediction, log10 y-scale, x pseudo-log thres = %s",
          thres
        )
      )
  }

  p
}
```

```{r}
thres <- 2

pred_df <- model_data |>
  make_prediction_grid(n = 1000) |>
  add_fixed_effect_predictions(model = m4)

plot_pred_raw <- plot_observed_vs_predicted(
  data = model_data,
  pred_df = pred_df,
  model_name = "m4",
  thres = thres,
  log_y = FALSE
)

plot_pred_log_y <- plot_observed_vs_predicted(
  data = model_data,
  pred_df = pred_df,
  model_name = "m4",
  thres = thres,
  log_y = TRUE
)

plot_pred_raw
plot_pred_log_y
```

Note:
Here we see that the predictions become non-physiological below ~200 mel EDI.
We will have to take care of this issue during the power analysis, more is
detailed on this in the relevant file: `power_analysis.Rmd`

# Create photopic and rhodopic models, too

This allows to assume that cones/rods are driving the responses, instead of just
assuming that ipRGCs are the only driver…

```{r}
#| fig-width: 8
#| fig-height: 12

form_m4_phot <- log_emg_raw ~
  log_illuminance +
  lights_on +
  log_baseline +
  trial +
  (log_illuminance | participant_id)

m4_phot <- fit_glmmTMB(form_m4_phot, model_data)

check_glmmTMB(m4_phot, save_path = here::here("outputs", "models", "emg_lmm_phot_diag.png"))
```

```{r}
#| fig-width: 8
#| fig-height: 12

form_m4_rho <- log_emg_raw ~
  log_rhodopic_edi +
  lights_on +
  log_baseline +
  trial +
  (log_illuminance | participant_id)

m4_rho <- fit_glmmTMB(form_m4_rho, model_data)

check_glmmTMB(m4_rho, save_path = here::here("outputs", "models", "emg_lmm_rho_diag.png"))
```

```{r}
AIC(m4, m4_phot, m4_rho)
```

As expected, the three models are pretty much equivalent, because we keep the
spectrum constant, the ratios between the metrics remain constant too

# Saving final models

```{r}
# Final models -------------------------------------------------------------

final_model_mel <- m4
final_model_phot <- m4_phot
final_model_rho <- m4_rho

final_models <- list(
  mel = final_model_mel,
  phot = final_model_phot,
  rho = final_model_rho
)

model_labels <- c(
  mel = "Melanopic EDI",
  phot = "Photopic illuminance",
  rho = "Rhodopic EDI"
)

light_metric_labels <- c(
  mel = "Melanopic EDI, log10",
  phot = "Photopic illuminance, log10",
  rho = "Rhodopic EDI, log10"
)

# Save final model objects -------------------------------------------------

purrr::iwalk(
  final_models,
  ~ save_rds_safe(
    .x,
    here::here(
      "outputs",
      "models",
      glue::glue("emg_lmm_{.y}.rds")
    )
  )
)
```

# Manuscript-ready outputs

```{r}
# Coefficient table --------------------------------------------------------

coef_table <- purrr::imap_dfr(
  final_models,
  ~ broom.mixed::tidy(
    .x,
    effects = "fixed",
    conf.int = TRUE
  ) %>%
    dplyr::mutate(
      model = .y,
      model_label = model_labels[[.y]],
      estimate_exp = exp(estimate),
      conf.low_exp = exp(conf.low),
      conf.high_exp = exp(conf.high)
    )
)

readr::write_csv(
  coef_table,
  here::here("outputs", "models", "emg_lmm_all_coefs.csv")
)

coef_table
```

```{r}
# Model summaries ---------------------------------------------------------

model_summaries <- purrr::imap_dfr(
  final_models,
  ~ {
    r2 <- MuMIn::r.squaredGLMM(.x)

    tibble::tibble(
      model = .y,
      model_label = model_labels[[.y]],
      r2_marginal = r2[1, "R2m"],
      r2_conditional = r2[1, "R2c"],
      AIC = AIC(.x),
      BIC = BIC(.x),
      logLik = as.numeric(logLik(.x))
    )
  }
)

readr::write_csv(
  model_summaries,
  here::here("outputs", "models", "emg_lmm_model_summaries.csv")
)

model_summaries
```

```{r, results='asis'}
# Save manuscript-style HTML model tables ---------------------------------

dir.create(
  here::here("outputs", "models"),
  recursive = TRUE,
  showWarnings = FALSE
)

sjPlot::tab_model(
  final_model_mel,
  show.icc = TRUE,
  show.r2 = TRUE,
  show.re.var = TRUE,
  pred.labels = c(
    "Intercept",
    "Melanopic EDI, log10",
    "Lights on",
    "Baseline EMG, log10",
    "Trial number"
  ),
  dv.labels = "OO-EMG response, log10",
  file = here::here("outputs", "models", "emg_lmm_mel_model_table.html")
)

sjPlot::tab_model(
  final_model_phot,
  show.icc = TRUE,
  show.r2 = TRUE,
  show.re.var = TRUE,
  pred.labels = c(
    "Intercept",
    "Photopic illuminance, log10",
    "Lights on",
    "Baseline EMG, log10",
    "Trial number"
  ),
  dv.labels = "OO-EMG response, log10",
  file = here::here("outputs", "models", "emg_lmm_phot_model_table.html")
)

sjPlot::tab_model(
  final_model_rho,
  show.icc = TRUE,
  show.r2 = TRUE,
  show.re.var = TRUE,
  pred.labels = c(
    "Intercept",
    "Rhodopic EDI, log10",
    "Lights on",
    "Baseline EMG, log10",
    "Trial number"
  ),
  dv.labels = "OO-EMG response, log10",
  file = here::here("outputs", "models", "emg_lmm_rho_model_table.html")
)
```

```{r}
sessionInfo()
```

```{r render-time-2, echo=FALSE}
render_time <- Sys.time() - .render_start
cat("Render time:", format(render_time))
```
